```{r, echo = FALSE}
title <- "expression"
```

---
title: `r title`
author: "Steffan Noe Niikanoff Christiansen"
output: html_document
---


### Global options and paths
```{r Global options and paths, echo = FALSE, message=FALSE}
# libraries
packages <- c("tidyverse", "patchwork", "kableExtra", "here", "knitr", "ggtranscript", "scales")

invisible(lapply(packages, require, character.only = TRUE)) 
theme_set(theme_bw() + 
            theme(plot.title = element_text(hjust = 0.5),
                  axis.text.x = element_text(angle = 45,
                                             margin = margin(4, 0, 0, 0),
                                             hjust = 1),
                  strip.text = element_text(size = 8),
                  strip.background = element_rect(fill="white"),
                  panel.spacing = unit(0.1, "lines")))

# paths
root_path <- here()
projects_path <- str_remove(root_path, "/[A-z]+$")
# tables_path <- here("results", "tables")
gtex_v9 <- here(projects_path, "gtex", "data", "gtex.v9")
gtex_v8 <- here(projects_path, "gtex", "data", "gtex.v8")
pdui_te_data <- here(projects_path, "public_transcriptomes", "results", "processed")
# gencode_filtered <- here(projects_path, "gtex", "results", "processed")
gtex_processed_path <- here(projects_path, "gtex", "results", "processed")


title <- str_replace_all(title, " ", "_")
fig_path <- here("results", "images", title, "/")
tab_path <- here("results", "tables", title, "/")

# global options
opts_knit$set(root.dir = root_path)
opts_chunk$set(message = FALSE,
               # dpi = 300,
               dpi = 1200,
               # dev = "jpeg",
               dev = "svg",
               fig.width = 6,
               fig.height = 6,
               fig.align="center", 
               fig.path = fig_path,
               echo = FALSE,
               warning = FALSE)

options(scipen = 100, digits = 3, bitmapType = "cairo") # fixes grid.Call warning 
time <- lubridate::date()

# helper functions
s_table <- function (df) {
    kableExtra::kable_styling(knitr::kable(df), bootstrap_options = c("condensed", 
        "striped"), full_width = FALSE, font_size = 14)
}

s_names <- function (df) 
{
    colnames(df) <- tolower(stringr::str_replace_all(colnames(df), 
        c(` \\| ` = "_", ` - ` = "_", ` ` = "_", `|` = "", `-` = "_", 
            `:` = "", `\\(` = "", `\\)` = "", `\\+` = "_", `#` = "")))
    return(df)
}

scroll_table <- function (dataframe) 
{
    kableExtra::scroll_box(kableExtra::kable_styling(knitr::kable(dataframe), 
        "striped", full_width = FALSE, font_size = 14), width = "1000px", 
        height = "500px")
}


# visulisation settings
colors <- read_csv(here(gtex_v8, "tissue_colors.csv"))

selected_colors <- list(
  gene = c(CALM1 = "#1B9E77",
           CALM2 = "#D95F02",
           CALM3 = "#7570B3"), 
  tissue = setNames(colors$color, colors$tissueSiteDetailId),
  measure = c("PDUI adjusted counts per nt" = "#A6761D", "Counts per nt" = "#666666"), 
  # type = c(five_prime_UTR = "#9E0142",
  #          CDS = "#FDAE61",
  #          three_prime_UTR = "#66C2A5"),
  type = c(five_prime_UTR = "#666666",
           CDS = "#66A61E",
           three_prime_UTR = "#A6761D"),
  MANE_status = c("MANE" = "red",
                  "non-MANE" = "grey")
  )

selected_theme_poster <- theme(
  axis.text.y = element_text(size = 18),
  axis.text.x = element_text(size = 18),
  axis.title.y = element_text(size = 20), 
  axis.title.x = element_text(size = 20), 
  legend.text = element_text(size = 18),
  legend.title = element_text(size = 20),
  title = element_text(size = 18))

selected_theme <- theme(axis.text.y = element_text(size = 14),
      axis.text.x = element_text(size = 14),
      axis.title.y = element_text(size = 16), 
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 14),
      strip.text = element_text(size = 12))
```

Rendering of the report was started `r time` <br> Following packages
were loaded: `r packages` <br><br>

### Gencode data
```{r}
gencode_v26_URL <- "https://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_human/release_26/gencode.v26.annotation.gff3.gz"
# tmp <- download.file(gencode_v26_URL, file.path(gtex_data_path, "gencode.v26"))
```
`gencode.v26` was downloaded from following URL: `r gencode_v26_URL`

```{r eval = FALSE}
gencode_v26 <- rtracklayer::import.gff3(here(gtex_v9, "gencode.v26"))

version <- gencode_v26
transcript_info <- function(symbol) {
  version %>% 
    plyranges::filter(gene_name == symbol) %>% 
    as_tibble() 
}

CALM1_v26 <- transcript_info("CALM1")
CALM2_v26 <- transcript_info("CALM2")
CALM3_v26 <- transcript_info("CALM3")

CALM1_unqiue_transcripts <- CALM1_v26$transcript_id %>% na.omit %>% unique 
CALM2_unqiue_transcripts <- CALM2_v26$transcript_id %>% na.omit %>% unique 
CALM3_unqiue_transcripts <- CALM3_v26$transcript_id %>% na.omit %>% unique 

CALM_V26_gencode <- 
  bind_rows(CALM1_v26, CALM2_v26, CALM3_v26) %>% 
  mutate(gencode_version = "v26")

# saveRDS(CALM_V26_gencode, "results/processed/gencode_v26_CALM1-3_2022.11.25.rds")
```

```{r}
CALM_V26_gencode <- readRDS(here(gtex_processed_path, "gencode_v26_CALM1-3_2022.11.25.rds"))

return_unique_transcripts <- function(gene) {
  CALM_V26_gencode %>% 
    filter(gene_name == gene) %>% 
    pull(transcript_id) %>% 
    na.omit %>% 
    unique 
}

CALM1_unqiue_transcripts <- return_unique_transcripts("CALM1")
CALM2_unqiue_transcripts <- return_unique_transcripts("CALM2")
CALM3_unqiue_transcripts <- return_unique_transcripts("CALM3")
```
<br><br>

#### Transcript lengths
```{r}
MANE_calm <- c("ENST00000356978", "ENST00000272298", "ENST00000291295")
CALM_V26_gencode %>% 
  filter(type == "exon") %>% 
  filter(str_detect(transcript_id, paste(MANE_calm, collapse = "|"))) %>% 
  group_by(gene_name, transcript_id) %>% 
  reframe(sum_of_exons_MANE = sum(width)) %>% 
  s_table()
```
<br><br>

#### Lengths of UTRs and CDS
```{r}
t0 <- CALM_V26_gencode %>% 
  filter(type == "five_prime_UTR") %>% 
  filter(str_detect(transcript_id, paste(MANE_calm, collapse = "|"))) %>% 
  group_by(gene_name) %>% 
  reframe("five_UTR" = sum(width))

t1 <- CALM_V26_gencode %>% 
  filter(type == "CDS") %>% 
  filter(str_detect(transcript_id, paste(MANE_calm, collapse = "|"))) %>% 
  group_by(gene_name) %>% 
  reframe(CDS = sum(width))

t2 <- CALM_V26_gencode %>% 
  filter(type == "three_prime_UTR") %>% 
  filter(str_detect(transcript_id, paste(MANE_calm, collapse = "|"))) %>% 
  group_by(gene_name) %>% 
  reframe("three_UTR" = sum(width))

t3 <- full_join(t0, t1) %>% 
  full_join(., t2, by = "gene_name") %>% 
  rowwise() %>% 
  mutate(total = sum(five_UTR + CDS + three_UTR),
         perc_three_UTR = three_UTR/total, five_UTR_CDS = five_UTR + CDS, 
         ratio_three_remaining = three_UTR/five_UTR_CDS)

colnames(t3) <- c("Gene", "5'-UTR", "CDS", "3'-UTR", "Total", "Fraction 3'-UTR", "5'-UTR + CDS", "3'-UTR/(5'-UTR + CDS)")
t3 %>% 
  s_table()
```

#### Transcript counts
```{r Transcript counts, fig.width = 10, fig.height = 6, out.width = "100%"}
CALM_V26_gencode %>%
  filter(type == "transcript") %>%
  dplyr::count(gene_name, gene_id) %>% 
  s_table()
```
<br><br>

### Visualised protein-coding transcripts
```{r Visualised protein-coding transcripts, fig.width = 10, fig.height = 6, out.width = "100%"}
# selected_colors <- list(
#   gene = c(CALM1 = "#1B9E77",
#            CALM2 = "#D95F02",
#            CALM3 = "#7570B3"),
#   type = c(five_prime_UTR,
#            CDS,
#            three_prime_UTR))

plot_calm <- function(selected_gene) {
  
  exons <- CALM_V26_gencode %>% 
    filter(type == "exon" & gene_name == selected_gene & 
             transcript_type == "protein_coding") 
  
  cds <- CALM_V26_gencode %>%
    dplyr::filter(type == "CDS" & gene_name == selected_gene & 
                    transcript_type == "protein_coding") 
  
  utrs <- CALM_V26_gencode %>% 
    filter(type %in% c("five_prime_UTR", "three_prime_UTR") & 
             gene_name == selected_gene & transcript_type == "protein_coding") 
  
  exons %>%
    ggplot(aes(xstart = start,
               xend = end,
               y = transcript_id)) +
    # geom_range(aes(fill = type),
    #            height = 0.25) +
    geom_range(data = cds,
               aes(fill = type)) +
    geom_range(data = utrs,
               aes(fill = type),
               height = 0.25) +
    geom_intron(data = to_intron(exons, "transcript_id"),
                aes(strand = strand),
                arrow.min.intron.length = arrow_treshold) +
    labs(x = "",
         # y = "Transcript ID",
         y = "",
         fill = "Transcript type") +
    scale_x_continuous(labels = comma) +
    # scale_fill_manual(values = selected_colors$type) +
    scale_fill_manual(values = selected_colors$type,
                    labels = c("CDS" = "CDS", 
                               "three_prime_UTR" = "3'-UTR", 
                               "five_prime_UTR" = "5'-UTR"))
}
arrow_treshold = 10000
```

#### CALM1 protein coding
```{r CALM1 proteincoding, fig.width = 8, fig.height = 3, out.width = "80%"}
plot_calm("CALM1") + 
  ggtitle(bquote(italic("CALM1"))) + 
  selected_theme
```
<br><br>

#### CALM2 protein coding
```{r CALM2 protein coding, fig.width = 8, fig.height = 3, out.width = "80%"}
plot_calm("CALM2") + 
  ggtitle(bquote(italic("CALM2"))) + 
  selected_theme
```
<br><br>

#### CALM3 protein coding
```{r CALM3 protein coding, fig.width = 8, fig.height = 3, out.width = "80%"}
plot_calm("CALM3") + 
  ggtitle(bquote(italic("CALM3"))) + 
  selected_theme
```
<br><br>

### GTEx v9 long-read data
```{r GTEx v9 long-read data, fig.width = 10, fig.height = 6, out.width = "100%"}
GTEx_v9_TPM_URL <- "https://storage.googleapis.com/gtex_analysis_v9/long_read_data/quantification_gencode.tpm.txt.gz"

GTEx_v9_TPM <- read_tsv(here(gtex_v9, "quantification_gencode.tpm.txt.gz"))
GTEx_v9_counts <- read_tsv(here(gtex_v9, "quantification_gencode.counts.txt.gz"))

CALM1_v9_TPM <- GTEx_v9_TPM %>% filter(transcript %in% CALM1_unqiue_transcripts)
CALM2_v9_TPM <- GTEx_v9_TPM %>% filter(transcript %in% CALM2_unqiue_transcripts)
CALM3_v9_TPM <- GTEx_v9_TPM %>% filter(transcript %in% CALM3_unqiue_transcripts)

CALM1_v9_counts <- GTEx_v9_counts %>% filter(transcript %in% CALM1_unqiue_transcripts)
CALM2_v9_counts <- GTEx_v9_counts %>% filter(transcript %in% CALM2_unqiue_transcripts)
CALM3_v9_counts <- GTEx_v9_counts %>% filter(transcript %in% CALM3_unqiue_transcripts)

sup_table1 <- readxl::read_xlsx(here(gtex_v9, "41586_2022_5035_MOESM4_ESM.xlsx"), sheet = 1) %>% 
  mutate(smtsd = tissue,
         tissue = str_replace_all(tissue, c("- " = "", " " = "_")) %>% 
           str_to_lower())
```
Long Read RNASeq data (TPM values) from GTEx (v9) was downloaded 2022.09.01 from `r GTEx_v9_TPM_URL`. Described in doi: 10.1038/s41586-022-05035-y (Nature 2022).

```{r}
rm(GTEx_v9_TPM, GTEx_v9_counts)
```

```{r}
df_to_tidy <- function(df, measure) {
  tidy_df <- df %>% 
    pivot_longer(!transcript, names_to = "sample_id", values_to = measure)
  tidy_df <- left_join(tidy_df, sup_table1) %>% 
    filter(!tissue %in% c("k562", "cells_cultured_fibroblasts"))
}

CALM_tidy_TPM <- bind_rows(
  df_to_tidy(CALM1_v9_TPM, "TPM") %>% mutate(gene = "CALM1"),
  df_to_tidy(CALM2_v9_TPM, "TPM") %>% mutate(gene = "CALM2"),
  df_to_tidy(CALM3_v9_TPM, "TPM") %>% mutate(gene = "CALM3")
) %>% 
  mutate(gene = as.factor(gene),
         MANE_status = if_else(
          str_detect(transcript, paste(MANE_calm, collapse = "|")),
          "MANE",
          "non-MANE"))

CALM_tidy_counts <- bind_rows(
  df_to_tidy(CALM1_v9_counts, "counts") %>% mutate(gene = "CALM1"),
  df_to_tidy(CALM2_v9_counts, "counts") %>% mutate(gene = "CALM2"),
  df_to_tidy(CALM3_v9_counts, "counts") %>% mutate(gene = "CALM3")
) %>% 
  mutate(gene = as.factor(gene),
         MANE_status = if_else(
           str_detect(transcript, paste(MANE_calm, collapse = "|")),
           "MANE",
           "non-MANE"))
```
K562 and cells_cultured_fibroblasts were removed. 
<br><br>

```{r adding total counts per sample}
total_TPM <- CALM_tidy_TPM %>% 
  group_by(sample_id) %>% 
  summarise(sum_TPM = sum(TPM))

CALM_tidy_TPM <- CALM_tidy_TPM %>% 
  left_join(., total_TPM, by = c("sample_id")) %>% 
  mutate(fraction_TPM = TPM/sum_TPM)

# CALM_tidy_TPM %>%
#   group_by(sample_id) %>%
#   reframe(total = sum(fraction_TPM)) %>%
#   count(total)

total_counts_v9 <- CALM_tidy_counts %>% 
  group_by(sample_id) %>% 
  summarise(sum_counts = sum(counts))

CALM_tidy_counts <- CALM_tidy_counts %>% 
  left_join(., total_counts_v9, by = c("sample_id")) %>% 
  mutate(rel_counts_v9 = counts/sum_counts)

# CALM_tidy_counts %>%
#   group_by(sample_id) %>%
#   reframe(total = sum(rel_counts_v9)) %>%
#   count(total)
```

#### Comparison of counts and TPM for estimation of fraction expressed
```{r Comparison of counts and TPM, fig.width = 10, fig.height = 6, out.width = "100%"}
df_fractions <- full_join(
  CALM_tidy_TPM %>% 
    select(sample_id, transcript, gene, tissue, fraction_TPM), 
  CALM_tidy_counts %>% 
    select(sample_id, transcript, gene, tissue, rel_counts_v9),
  by = c("sample_id", "transcript", "gene", "tissue"))

df_fractions %>% 
  ggplot(aes(x = fraction_TPM, y = rel_counts_v9, color = tissue)) + 
  facet_wrap(~gene, ncol = 3) + 
      geom_abline(intercept = 0, slope = 1, color = "darkgrey", linetype= "dashed", linewidth = 0.75) +
  geom_point() + 
  coord_fixed() 
```
<br><br>

#### MANE expression per gene and tissue (among protein coding)
```{r}
protein_coding_transcripts <- CALM_V26_gencode %>% 
  filter(transcript_type == "protein_coding") %>% 
  pull(transcript_id) %>% 
  unique()

CALM_tidy_TPM <- CALM_tidy_TPM %>% 
  filter(transcript %in% protein_coding_transcripts)
```

```{r}
total_TPM <- CALM_tidy_TPM %>% 
  group_by(sample_id, gene) %>% 
  summarise(sum_TPM = sum(TPM))

CALM_tidy_TPM <- CALM_tidy_TPM %>% 
  select(-c(fraction_TPM, sum_TPM)) %>% 
  left_join(., total_TPM, by = c("sample_id", "gene")) %>% 
  mutate(fraction_TPM = TPM/sum_TPM)

CALM_tidy_TPM %>%
  select(sample_id, tissue, gene, transcript, TPM, sum_TPM, fraction_TPM, MANE_status) %>% 
  write_csv(., here(tab_path, "relative_expression_per_transcript.csv"))

t1 <- CALM_tidy_TPM %>%
  filter(MANE_status == "MANE") %>%
  select(sample_id, tissue, gene, transcript, fraction_TPM, MANE_status) %>%
  group_by(gene, tissue) %>%
  reframe(min = min(fraction_TPM),
          median = median(fraction_TPM),
          max = max(fraction_TPM))
t1 %>% 
  dplyr::select(gene, tissue, median) %>% 
  pivot_wider(names_from = gene, values_from = median) %>% 
  s_table()
```

#### MANE per gene
```{r MANE expression per gene and tissue, fig.width = 10, fig.height = 6, out.width = "100%"}
CALM_tidy_TPM %>%
  filter(MANE_status == "MANE") %>%
  select(sample_id, tissue, gene, transcript, fraction_TPM, MANE_status) %>%
  group_by(gene) %>%
  reframe(min = min(fraction_TPM),
          median = median(fraction_TPM),
          max = max(fraction_TPM)) %>% 
  s_table()
```
<br>

#### CALM1
```{r}
t1_CALM1 <- t1 %>% 
  filter(gene == "CALM1")

t1_CALM1 %>% s_table()
```
Median of medians is `r median(t1_CALM1$median)` (range: `r min(t1_CALM1$min)`-`r max(t1_CALM1$max)`) <br>

#### CALM2
```{r}
t1_CALM2 <- t1 %>% 
  filter(gene == "CALM2")

t1_CALM2 %>% s_table()
```
Median of medians is `r median(t1_CALM2$median)` (range: `r min(t1_CALM2$min)`-`r max(t1_CALM2$max)`) <br>

#### CALM3
```{r}
t1_CALM3 <- t1 %>% 
  filter(gene == "CALM3")

t1_CALM3 %>% s_table()
```
Median of medians is `r median(t1_CALM3$median)` (range: `r min(t1_CALM3$min)`-`r max(t1_CALM3$max)`) <br><br>

#### MANE cardiac tissue
```{r MANE cardiac tissue 8_4, fig.width = 8, fig.height = 3, out.width = "100%"}
plot_mane_fraction <- function(gene_name){
  MANE_transcript <- CALM_tidy_TPM %>% 
    filter(gene == gene_name) %>% 
    filter(MANE_status == "MANE") %>% 
    pull(transcript) %>% 
    unique()

    temp_df <- CALM_tidy_TPM %>%
    filter(tissue  == "heart_left_ventricle") %>% 
    filter(transcript == MANE_transcript) %>%
    select(sample_id, tissue, gene, transcript, fraction_TPM, MANE_status) %>%
    group_by(sample_id, tissue, gene, MANE_status)
  
  temp_df %>%
    ggplot(aes(y = fraction_TPM, x = transcript)) +
    # ggplot(aes(y = fraction_TPM, x = transcript, color = MANE_status)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.2) +
    geom_jitter(width = 0.2, size = 2.5, alpha = 1, height = 0) +
    theme(axis.text.x = element_text(size = 12, angle = 0, hjust = 0.5),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_text(size = 16),
        axis.title.y = element_text(size = 16),
        legend.title = element_text(size = 16),
        legend.text = element_text(size = 14),
        panel.spacing = unit(0.5, "cm")) +
  theme(legend.position = "none") +
    # scale_color_manual(values = selected_colors$MANE_status) +
    labs(
      x = "",
      title = bquote(~italic(.(gene_name))),
      # y = "Percentage expressed \nby MANE transcripts",
      y = "Relative expression")
}
p1 <- plot_mane_fraction("CALM1") 
p2 <- plot_mane_fraction("CALM2") 
p3 <- plot_mane_fraction("CALM3") 

p_comb <- p1 + p2 + p3 + plot_layout(ncol = 3, guides = "collect", axes = "collect") &
  scale_y_continuous(
    limits = c(0.975, 1),
                     breaks = seq(0.975, 1, by = 0.005),
                     labels = scales::percent_format())
p_comb
```

```{r MANE cardiac tissue 7_3, fig.width = 7, fig.height = 3, out.width = "100%"}
p_comb
```

<br><br>

### Investigation of PDUI values
PDUI: percentage of distal poly(A) site usage index (PDUI) value <br>

```{r eval = FALSE}
untar("data/PDUI/To_CALM1.tar.gz", exdir = "data/PDUI")
```

```{r}
pdui_dir <- here(projects_path, "gtex/data/PDUI/To_CALM1")

calm_transcript_id <- c("NM_001363670", "NM_006888", "NM_001743", "NM_001305626", "NM_001329922")

cam_genes <- c("CALM1", "CALM2", "CALM3")

combined_pdui <- tibble()
for (tissue in dir(pdui_dir)){
  temp_df <- read_tsv(file.path(pdui_dir, tissue)) %>% 
    s_names
  
  tissue_name <- str_remove(tissue, "_All_PDUIs.txt.gz")
  colnames(temp_df)[1] <- "name"
  
  temp_df <- temp_df %>% 
    filter(str_detect(name, paste(calm_transcript_id, collapse = "|"))) %>% 
    mutate(gene = str_extract(name, "CALM[1-3]") %>% factor(levels = cam_genes),
           trans_id = str_extract(name, "NM_[0-9]+") %>% factor()) %>% 
    dplyr::select(gene, trans_id, everything())
  
  temp_df <- temp_df %>% 
    pivot_longer(cols = -c(gene:loci), names_to = "sample_name", values_to = "pdui") %>% 
    mutate(tissue = tissue_name)
  
  combined_pdui <- bind_rows(temp_df, combined_pdui)
}

tissue_count <- combined_pdui %>% 
  dplyr::count(tissue) %>% 
  mutate(samples_per_tissue = n/length(calm_transcript_id)) %>% 
  dplyr::select(-n)

sample_count <- combined_pdui %>% 
  dplyr::count(sample_name) %>% 
  mutate(samples_per_id = n/length(calm_transcript_id)) %>% 
  dplyr::select(-n)

sample_brain_count <- combined_pdui %>% 
  mutate(brain_sample = str_detect(tissue, "Brain")) %>% 
  dplyr::count(sample_name, brain_sample) %>%
  filter(brain_sample == TRUE) %>% 
  mutate(brain_samples_per_id = n/length(calm_transcript_id)) %>% 
  dplyr::select(-c(brain_sample, n))

sample_count <- left_join(sample_count,
                          sample_brain_count) %>% 
  mutate(brain_samples_per_id = 
           if_else(is.na(brain_samples_per_id), 0, brain_samples_per_id))

rm(sample_brain_count)
```
In accordance with 2021 NAR paper, there are data from `r combined_pdui %>% dplyr::count(sample_name, tissue) %>% nrow()` samples and `r nrow(sample_count)` individuals. `r nrow(tissue_count)` tissues are represented. The number of brain samples per individual ranges from `r min(sample_count$brain_samples_per_id)` to `r max(sample_count$brain_samples_per_id)` (n individuals with brain samples: `r sample_count %>% filter(brain_samples_per_id > 0) %>% nrow()`). <br>

Data are available for following tissues:
```{r}
tissue_count %>% scroll_table()
```
<br>

Data are available for following loci:
```{r}
t1 <- combined_pdui %>% 
  arrange(gene, trans_id) %>% 
  distinct(gene, trans_id, loci) %>% 
  mutate(start = str_remove(loci, "chr[0-9]+:") %>% 
           str_extract(., "^[0-9]+") %>% as.numeric,
         end = str_extract(loci, "[0-9]+$") %>% as.numeric,
         width = end-start)

t1 %>% s_table()
```
Notice: <br> 
  
  * CALM1 start for both transcripts are identical with exon 9 start for gencode collapsed model (v26). CALM1 end of NM_001363670 and NM_006888 are 14 and 7 nuclotides away from CALM1 end for gencode collapsed model (v26). 
  * CALM2 start are identical or two nt away from start of exon 8 for gencode collapsed model (v26). CALM2 end are identical with end of exon 8 for gencode collapsed model. 
  * CALM3 start are identical with start of exon 9 for gencode collapsed model (v26). CALM3 end are 11 nucleotides away from end of exon 11 for gencode collapsed model. 
<br><br>

  
#### Correlation among transcripts
```{r Correlation among transcripts, fig.width = 10, fig.height = 5, out.width = "80%"}
x_limits <- c(0, 1)
y_limits <- c(0, 1)

return_wide <- function(long_df, trans_id1, trans_id2) {
  wide_df <- long_df %>% 
    filter(trans_id == trans_id1 | trans_id == trans_id2) %>%
    dplyr::select(-c(name,loci)) %>% 
    pivot_wider(names_from = trans_id, values_from = pdui) %>% 
    mutate(difference = get(paste(trans_id1))-get(paste(trans_id2)))
}

# plot_scatter <- function(df, trans_id1, trans_id2) {
#   temp_df <- return_wide(df, trans_id1, trans_id2)
#   
#   gene_name <- temp_df %>%
#     pull(gene) %>%
#     unique
# 
#   temp_df %>%
#     ggplot(., aes(x = get(paste(trans_id1)), 
#                   y = get(paste(trans_id2)), 
#                   color = gene)) +
#     geom_point() +
#     theme(legend.position = "none") + 
#     scale_color_manual(values = selected_colors$gene) +
#     geom_abline(intercept = 0, slope = 1, color = "darkgrey", linetype= "dashed", linewidth = 0.75) +
#     scale_x_continuous(limits = x_limits) +
#     scale_y_continuous(limits = y_limits) +
#     labs(title = gene_name, 
#          x = trans_id1,
#          y = trans_id2) +
#     coord_fixed()
# }
# 
# p1 <- plot_scatter(combined_pdui, "NM_001363670", "NM_006888")
# p2 <- plot_scatter(combined_pdui, "NM_001743", "NM_001305626")
# 
# p1 + p2
```
<br><br>

```{r}
CALM1_wide <- return_wide(combined_pdui, "NM_001363670", "NM_006888")
CALM2_wide <- return_wide(combined_pdui, "NM_001743", "NM_001305626")
```
Two transcripts are detected for CALM1 and CALM2. The max difference between the two transcripts are `r max(CALM1_wide$difference)` and `r max(CALM2_wide$difference)` for CALM1 and CALM2, respectively. The pearson correlation between the two transcripts are `r cor(CALM1_wide$NM_001363670, CALM1_wide$NM_006888, method = "pearson")` and `r cor(CALM2_wide$NM_001743, CALM2_wide$NM_001305626, method = "pearson")` for CALM1 and CALM2, respectively. 
<br>


```{r}
combined_pdui <- combined_pdui %>% 
  group_by(gene, sample_name, tissue) %>% 
  reframe(pdui = median(pdui))
```
Due to the similarity the analyses are conducted based on the median between the two transcripts. 


#### All tissues PDUI
```{r All tissues PDUI, fig.width = 8, fig.height = 8, out.width = "100%"}
plot_pdui_distribution <- function(selected_gene) {
  combined_pdui %>% 
    filter(gene == selected_gene) %>% 
    ggplot(., aes(x = tissue, y = pdui, color = tissue)) +
    geom_boxplot(outlier.shape = NA) + 
    geom_jitter(width = 0.05, size = 0.15, alpha = .25, height = 0) +
    theme(legend.position = "none", 
          plot.margin = unit(c(0.2, 0.2, 0.2, 1.5), "cm"),
          plot.title = element_text(size = 12),
          strip.text = element_text(face = "italic"),
          axis.text.y = element_text(size = 14),
        axis.title.y = element_text(size = 16)) +
    scale_color_manual(values = selected_colors$tissue) + 
    labs(title = bquote(~italic(.(selected_gene))),
         y = "Percentage of distal poly(A) site usage index (PDUI)",
         x = "")
}

for (i in 1:3) {
  temp_plot <- plot_pdui_distribution(cam_genes[i])
  assign(paste0("p", i), temp_plot, envir = .GlobalEnv)
}

p1 + p2 + p3 +
  plot_layout(ncol = 1, guides = "collect", axes = "collect") & 
  scale_y_continuous(limits = c(0,1.01)) &
  scale_x_discrete(labels = function(x) str_replace_all(x, "_", " ")) 
```
<br><br>
  
```{r}
combined_pdui %>% 
  group_by(gene) %>% 
  reframe(min_pdui = min(pdui),
          median_pdui = median(pdui),
          max_pdui = max(pdui),
          range_pdui = max_pdui-min_pdui,
          sd_pdui = sd(pdui)) %>% 
  s_table()
```
<br>


#### LV PDUI 
```{r }
LV_pdui <- combined_pdui %>% 
  filter(tissue == "Heart_Left_Ventricle")
```
Data are available for `r LV_pdui %>% distinct(sample_name) %>% nrow()` LV samples.

```{r boxplot pdui lv, fig.width = 4, fig.height = 4, out.width = "80%"}
p1 <- LV_pdui %>% 
  ggplot(., aes(x = gene, y = pdui, color = gene)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.35, size = 1, alpha = .15, height = 0) +
  theme(legend.position = "none",
        axis.text.x = element_text(
          face = "italic", 
          angle = 0,
          margin = margin(4, 0, 0, 0),
          hjust = 0.5,
          size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.y = element_text(size = 16),
        axis.title.x = element_text(size = 16)) + 
  scale_color_manual(values = selected_colors$gene) +
  labs(
    y = "Percentage of distal poly(A) \n site usage index (PDUI)",
       # y = "PDUI",
       x = "") 
p1 
```

```{r boxplot pdui lv 5_4, fig.width = 5, fig.height = 4, out.width = "80%"}
p1
```


```{r}
LV_pdui %>% 
  group_by(gene) %>% 
  reframe(min_pdui = min(pdui),
          median_pdui = median(pdui),
          max_pdui = max(pdui)) %>% 
  s_table
```
<br><br>

### PDUI weighted normalisation
```{r}
reads_cam <- readRDS(here(gtex_processed_path, "reads_cam_2024_08_21.rds"))
reads_cam <- left_join(reads_cam, colors) %>%
  mutate(sample_name = tolower(subjid) %>% str_replace(., "-", "_"))

colnames(reads_cam) <- str_replace_all(
  colnames(reads_cam),
  c("gene_symbol" = "gene", "tissueSiteDetailId" = "tissue"))

tpm_cam <- readRDS(here(gtex_processed_path, "tpm_cam_2024_08_21.rds"))
tpm_cam <- left_join(tpm_cam, colors) %>%
  mutate(sample_name = tolower(subjid) %>% str_replace(., "-", "_"))

colnames(tpm_cam) <- str_replace_all(
  colnames(tpm_cam),
  c("gene_symbol" = "gene", "tissueSiteDetailId" = "tissue"))

combined_pdui <- inner_join(
  combined_pdui,
  tpm_cam %>%
    dplyr::select(c("gene", "tissue", "sample_name", "value")),
  by = c("gene", "tissue", "sample_name"))

combined_pdui <- rename(combined_pdui, TPM = value)

median_counts <- tibble(
  gene = c("CALM1", "CALM2", "CALM3"),
  five_prime = c(248, 158, 199),
  cds = c(450), 
  three_prime = c(3544, 694, 1640),
  full_length = five_prime + cds + three_prime,
  pdui = c(0.39, 0.805, 0.64),
  counts = c(20603, 13908, 10502), 
  counts_per_nt = counts/full_length,
  counts_per_nt_pdui_adj = counts/(cds + five_prime + (three_prime*pdui)))

combined_pdui <- inner_join(
  combined_pdui,
  reads_cam,
  by = c("gene", "tissue", "sample_name")) %>%
  left_join(., median_counts %>% dplyr::select(gene:full_length)) %>%
  mutate(counts_per_nt = value/full_length,
         counts_per_nt_pdui_adj = value/(cds + five_prime + (three_prime*pdui)))

combined_pdui <- rename(combined_pdui, reads = value)
```
TPM were calculated as described in [Zhao et al](https://translational-medicine.biomedcentral.com/articles/10.1186/s12967-021-02936-w). The ratio between counts and transcript length and sum of all counts per transcripts are multiplied with $10^6$. In this proposed weighted normalisation, the denominator is superficial since we want to compare from gene to gene among each individual. The normalisation is conducted by dividing the counts with a length of the **used** transcript rather than length of the collapsed gene length: <br> 

$$
\frac{\text{counts}}{\text{cds_length} + \text{five_prime_length} + (\text{three_prime_length} \cdot \text{pdui})}
$$
<br><br>
  
#### Example 1: Identical counts and pdui equal to 1
```{r PDUI weighted normalisation, fig.width = 10, fig.height = 6, out.width = "100%"}
tibble(five_prime = c(50),
       cds = c(450), 
       three_prime = c(500, 1000, 3000),
       full_length = five_prime + cds + three_prime,
       pdui = c(1),
       counts = c(10000), 
       counts_per_nt = counts/full_length,
       counts_per_nt_pdui_adj = counts/(cds + five_prime + (three_prime*pdui))) %>% 
  s_table()
```
<br>

#### Example 2: Identical counts and pdui equal to 0
```{r}
tibble(five_prime = c(50),
       cds = c(450), 
       three_prime = c(500, 1000, 3000),
       full_length = five_prime + cds + three_prime,
       pdui = c(0),
       counts = c(10000), 
       counts_per_nt = counts/full_length,
       counts_per_nt_pdui_adj = counts/(cds + five_prime + (three_prime*pdui))) %>% 
  s_table()
```
<br>

#### LV median expression values
```{r}
median_counts %>% 
  s_table()
```
Median counts are obtained from **Summaries of read counts** (file:///C:/Users/QJ17ZB/R/projects/gtex/reports/CALM1-3_lv_v8_2024_04_05.html). Gene lengths correspond to those from **Gencode v26**.
<br><br>

#### Example 3: *CALM1*
```{r }
t1 <- median_counts %>% 
  filter(gene == "CALM1") %>% 
  dplyr::select(gene:full_length)

t2 <- bind_rows(t1, t1) %>% 
  mutate(pdui = c(1, 0), 
         counts = 10000, 
         counts_per_nt = counts/full_length, 
         counts_per_nt_pdui_adj = counts/(cds + five_prime + (three_prime*pdui))) 

t3 <- t2 %>% 
  dplyr::select(-gene) %>% 
  mutate(" " = c("#1", "#2"),
         across(everything(), ~format(., big.mark = ","))) %>% 
  dplyr::select(" ", everything())


colnames(t3) <- str_replace_all(colnames(t3), c("_" = " "))
colnames(t3)[2:9] <- c(
  "5'-UTR", "CDS", "3'-UTR", "Total length", "PDUI", 
  "Counts", "Counts per nt", "PDUI adjusted counts per nt")

t3 %>% 
  kable(align = "r") %>% 
  kableExtra::kable_styling(
    bootstrap_options = c("condensed", "striped"), 
    full_width = FALSE, 
    font_size = 14)
```
There is a `r t2$counts_per_nt_pdui_adj[2]/t2$counts_per_nt_pdui_adj[1]`-fold increase for #1 compared to #2. 
  
  
### Validation of approach using v8 and v9 GTEx data
```{r Validation of approach using v8 and v9 GTEx data, fig.width = 10, fig.height = 4, out.width = "100%"}
combined_pdui_adj <- combined_pdui %>%
  dplyr::select(gene, sampid, sample_name, tissue, smtsd, pdui, counts_per_nt, counts_per_nt_pdui_adj)

total_counts_v8 <- combined_pdui %>% 
  group_by(sampid) %>% 
  reframe(total_counts = sum(counts_per_nt),
          total_counts_pdui_adj = sum(counts_per_nt_pdui_adj))

combined_pdui_adj <- left_join(combined_pdui_adj, total_counts_v8, by = "sampid") %>% 
  mutate(rel_count = counts_per_nt/total_counts,
         rel_count_pdui_adj = counts_per_nt_pdui_adj/total_counts_pdui_adj)

# Counts summarises to one for each sample as expected
# combined_pdui_adj %>% 
#   group_by(sampid) %>%
#   reframe(sum_rel_count = sum(rel_count),
#           sum_rel_count_pdui_adj = sum(rel_count_pdui_adj)) %>% 
#   count(sum_rel_count, sum_rel_count_pdui_adj)

combined_v8_v9 <- inner_join(
  CALM_tidy_counts %>%
    filter(MANE_status == "MANE") %>% 
    select(sample_name, gene, smtsd, rel_counts_v9, sample_id) %>% 
    mutate(sample_name = tolower(sample_name) %>% 
             str_replace(., "-", "_")),
  combined_pdui_adj,
  by = c("sample_name", "smtsd", "gene"))

# combined_v8_v9 %>% dplyr::count(smtsd)

combined_v8_v9_long <- combined_v8_v9 %>% 
  pivot_longer(
    cols = c(rel_count, rel_count_pdui_adj), 
    values_to = "rel_counts_v8", 
    names_to = "measure") %>% 
  mutate(measure = str_replace_all(
    measure, 
    c("rel_count_pdui_adj" = "PDUI adjusted counts per nt",
      "rel_count" = "Counts per nt")))

p1 <- combined_v8_v9_long %>% 
  ggplot(aes(x = rel_counts_v8, y = rel_counts_v9, color = measure)) + 
  geom_point(alpha = 0.5) + 
  facet_grid(~gene) +
  theme(axis.text.x = element_text(size = 14, angle = 0, hjust = 0.5),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_text(size = 16),
        axis.title.y = element_text(size = 16),
        strip.text = element_text(size = 14, face = "italic"),
        legend.title = element_text(size = 16),
        legend.text = element_text(size = 14),
        panel.spacing = unit(0.5, "cm")) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, .5)) +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, .5)) +
  scale_color_manual(values = selected_colors$measure,
                     labels = function(x) str_replace_all(x, "counts", "\ncounts")) + 
  coord_fixed() + 
  geom_abline(slope = 1, linetype = "dashed", linewidth = 1) +
  labs(color = "Approach", 
       x = "Relative gene expression (v8)",
       y = "Relative gene expression (v9)") +
  scale_y_continuous(
    labels = scales::label_percent(), 
    breaks = seq(0, 1, by = 0.50), 
    limits = c(0,1)) +
  scale_x_continuous(
    labels = scales::label_percent(), 
    breaks = seq(0, 1, by = 0.50), 
    limits = c(0,1))
```

`r length(unique(combined_v8_v9_long$sample_id))` samples were available for the comparison of short-sequencing and long-read sequencing data. 
<br><br>

```{r Validation of approach using v8 and v9 GTEx data 2,fig.width = 6, fig.height = 7, out.width = "80%"}
p2 <- p1 + facet_wrap(~gene, ncol = 1) + theme(panel.spacing = unit(0.1, "cm"))
p2 
```
<br><br>
  

#### Distance to identity line
```{r Distance to identity line, fig.width = 10, fig.height = 4, out.width = "100%"}
distance_to_identity <- function(x, y) {
  abs(y - x) / sqrt(2)
}

# distance_to_identity(0.3, 0.3)
# distance_to_identity(0.3, 0)
# sqrt(0.212^2 + 0.212^2)
# distance_to_identity(0.9, 0.5)
# sqrt(0.283^2 + 0.283^2)
# distance_to_identity(1, 0)
# sqrt(0.707^2 + 0.707^2)
# distances to identity line as expected

combined_v8_v9_long <- combined_v8_v9_long %>% 
  mutate(distance = distance_to_identity(rel_counts_v8, rel_counts_v9)*100)

p3 <- combined_v8_v9_long %>% 
  ggplot(aes(x = measure, y = distance)) +
  geom_boxplot(outlier.shape = NA) + 
  geom_jitter(aes(color = measure), width = 0.20, alpha = .75, size = 1.5, height = 0) +
  scale_color_manual(values = selected_colors$measure, 
                     labels = function(x) str_replace_all(x, "counts", "\ncounts")) + 
  facet_wrap(~gene) + 
  theme(axis.text.x = element_text(size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_text(size = 16),
        axis.title.y = element_text(size = 16),
        strip.text = element_text(size = 14, face = "italic"),
        legend.title = element_text(size = 16),
        legend.text = element_text(size = 14)) + 
  labs(y = "Distance to identity line", 
       x = "", 
       color = "Approach") 
# p3
```
<br><br>

```{r Distance to identity line 2, fig.width = 6, fig.height = 7, out.width = "80%"}
p4 <- p3 + scale_x_discrete(labels = function(x) str_replace_all(x, "adjusted", "adjusted\n")) + facet_wrap(~gene, ncol = 1) + labs(x = "")

# p4 + theme(axis.text.x = element_text(size = 14, angle = 0, hjust = 0.5))
```
<br>

#### Validation and distance to identity line
```{r Validation and distance to identity line 10_6, fig.width = 10, fig.height = 6, out.width = "100%"}
p2 + theme(legend.position = "none") + 
  p4 +
  theme(axis.title.x = element_blank()) + 
  theme(axis.text.x = element_text(size = 14, angle = 0, hjust = 0.5)) +
  plot_layout(ncol = 2, guides = "collect")  + 
# &
#   theme(panel.spacing = unit(0.1, "lines"),
        # plot.margin = unit(c(0.1, 0.5, 0.1, 0.1), "cm")) +
  theme(strip.text = element_text(size = 10)) 
```
<br><br>

##### Median values and percent decreases
```{r}
combined_v8_v9_long %>% 
  group_by(gene, measure) %>% 
  reframe(median_distance = median(distance)) %>%
  pivot_wider(names_from = measure, values_from = median_distance) %>% 
  mutate(percentage_point_difference = `Counts per nt`-`PDUI adjusted counts per nt`,
         percent_decrease = percentage_point_difference/`Counts per nt`*100) %>% 
  s_table()
```
<br><br>


#### Significance testing
```{r}
distances_nested <- combined_v8_v9_long %>% 
  select(sample_name, sample_id, tissue, gene, measure, distance) %>% 
  mutate(distance = as.numeric(distance)) %>% 
  pivot_wider(names_from = measure, values_from = distance) %>% 
  group_by(gene) %>% 
  nest()

# shapiro <- function(data) {
#   test_result1 <- shapiro_counts_per_nt <- shapiro.test(data$`Counts per nt`)
#   test_result2 <- shapiro_counts_per_nt <- shapiro.test(data$`PDUI adjusted counts per nt`)
#   df1 <- broom::tidy(test_result1) 
#   df2 <- broom::tidy(test_result2) 
# }

# distances_nested %>%
#   mutate(test_results1 = map(data, shapiro),
#          test_results2 = map(data, shapiro)) %>%
#   unnest(test_results1, test_results2) %>%
#   ungroup %>% 
#   select(contains(c("p.value", "method")))
# Assumptions for t.test violated. 

wilcoxon_signed_ranked <- function(data) {
  test_result <- wilcox.test(
        data$`Counts per nt`,
        data$`PDUI adjusted counts per nt`,
        paired = TRUE, 
        alternative = "two.sided")
      broom::tidy(test_result)
}

wilcoxon_signed_ranked_p_values <- distances_nested %>%
  mutate(test_results = map(data, wilcoxon_signed_ranked)) %>%
  unnest(test_results) %>%
  ungroup

wilcoxon_signed_ranked_p_values %>%
  select(gene, p.value, statistic, method) %>% 
  mutate(p.value = sprintf("%.2e", p.value)) %>% 
  s_table()
```
<br><br>
  
  
#### LV expression
```{r LV expression, fig.width = 8, fig.height = 5, out.width = "100%"}
combined_pdui_LV <- combined_pdui %>% 
  filter(str_detect(tissue, "Left")) %>% 
  dplyr::select(gene, sample_name, counts_per_nt, counts_per_nt_pdui_adj) %>% 
  arrange(sample_name) 

total_counts <- combined_pdui %>% 
  filter(str_detect(tissue, "Left")) %>% 
  arrange(sample_name) %>% 
  group_by(sample_name) %>% 
  reframe(total_counts = sum(counts_per_nt),
          total_counts_pdui_adj = sum(counts_per_nt_pdui_adj))

combined_pdui_LV <- left_join(combined_pdui_LV, total_counts, by = "sample_name") %>% 
  mutate(rel_count = counts_per_nt/total_counts,
         rel_count_pdui_adj = counts_per_nt_pdui_adj/total_counts_pdui_adj)

# combined_pdui_LV %>%  
#   group_by(sample_name) %>%
#   reframe(sum(rel_count),
#           sum(rel_count_pdui_adj))

combined_pdui_LV_long <- combined_pdui_LV %>% 
  dplyr::select(-c(counts_per_nt:total_counts_pdui_adj)) %>% 
  pivot_longer(cols = rel_count:rel_count_pdui_adj, names_to = "measure", values_to = "relative_count") %>% 
  mutate(measure = str_replace_all(
    measure, 
    c("rel_count_pdui_adj" = "PDUI adjusted counts per nt",
      "rel_count" = "Counts per nt")))

p3 <-  combined_pdui_LV_long %>% 
  ggplot(., aes(x = gene, y = relative_count, color = gene)) +
  geom_violin(draw_quantiles = c(0.5)) +
  geom_jitter(width = 0.20, size = 0.50, alpha = 0.5, height = 0) +
  scale_color_manual(values = selected_colors$gene) +
  theme(legend.text = element_text(face = "italic"),
        axis.text.x = element_text(face = "italic", size = 14, angle = 0, hjust = 0.5),
        axis.text.y = element_text(size = 14),
        axis.title.y = element_text(size = 16)) + 
  facet_wrap(~measure) + 
  scale_y_continuous(limits = c(0,1)) + 
  labs(y = "Fraction of read counts per individual",
       x = "", 
       color = "Gene name") + 
  theme(legend.position = "none") 

p3 + theme(strip.text = element_text(size = 14))
```

```{r LV expression w4_h6, fig.width = 4, fig.height = 4, out.width = "80%"}
combined_pdui_LV_long %>% 
  filter(measure == "PDUI adjusted counts per nt") %>% 
  ggplot(., aes(x = gene, y = relative_count, color = gene)) +
  # geom_boxplot(outlier.shape = NA) +
    geom_violin(draw_quantiles = c(0.5)) +
  geom_jitter(width = 0.35, size = 1, alpha = 0.15, height = 0) +
  scale_color_manual(values = selected_colors$gene) +
  theme(legend.text = element_text(face = "italic"),
        axis.text.x = element_text(face = "italic", size = 14, angle = 0, hjust = 0.5),
        axis.text.y = element_text(size = 14),
        axis.title.y = element_text(size = 16),
        axis.title.x = element_text(size = 16)) + 
  scale_y_continuous(expand = c(0, 0),
                     breaks = seq(0, 1, by = 0.2), 
                     labels = scales::percent_format(),
                     limits = c(0,0.7)) + 
  labs(y = "Relative gene expression",
       x = "") + 
  theme(legend.position = "none") 
```

```{r}
t1 <- combined_pdui_LV_long %>% 
  group_by(gene, measure) %>% 
  reframe(median_relative_count = median(relative_count),
          mean_relative_count = mean(relative_count)) %>% 
  arrange(measure) 

CALM1_CALM2_diff <-  t1$median_relative_count[4]-t1$median_relative_count[5]

CALM1_CALM3_diff <- t1$median_relative_count[4]-t1$median_relative_count[6]

CALM2_CALM3_diff <- t1$median_relative_count[5]-t1$median_relative_count[6]

t1 %>% 
  s_table
```
<br><br>
  
  * CALM1-CALM2: `r CALM1_CALM2_diff`
  * CALM1-CALM3: `r CALM1_CALM3_diff`
  * CALM2-CALM3: `r CALM2_CALM3_diff`
  
#### Permutation test LV
```{r Permutation test LV, fig.width = 10, fig.height = 6, out.width = "100%", eval = FALSE}
median_df <- tibble()
# n <- 10
n <- 10000

return_diff <- function(input_df, gene_a, gene_b) {
  median_a <- input_df %>%
    filter(gene == gene_a) %>%
    reframe(median = median(shuffled)) %>%
    pull()
  median_b <- input_df %>%
    filter(gene == gene_b) %>%
    reframe(median = median(shuffled)) %>%
    pull()
  
  diff <- median_a-median_b
  return(diff)
}

permutation_test_pairwise <- function(
    input_df, variable_of_interest, gene_a, gene_b, n_iterations) {
  
  input_df <- input_df %>%
    filter(gene %in% c(gene_a, gene_b))
  
  for (i in seq(1:n_iterations)) {
    temp_df <- input_df %>%
      group_by(sample_name) %>%
      
      mutate(shuffled = sample(get(paste(variable_of_interest)), replace = FALSE),
             iteration = i) %>% 
      # mutate(shuffled = sample(relative_count, replace = FALSE),
      #        iteration = i) %>%
      ungroup()
    
    median_difference <- tibble(diff = return_diff(temp_df, gene_a, gene_b),
                                iteration = i,
                                gene_a = gene_a,
                                gene_b = gene_b)
    
    median_df <- bind_rows(median_df, median_difference)
  }
  return(median_df)
}

set.seed(1)
df1 <- permutation_test_pairwise(combined_pdui_LV, "rel_count_pdui_adj", "CALM1", "CALM2", n)
df2 <- permutation_test_pairwise(combined_pdui_LV, "rel_count_pdui_adj", "CALM1", "CALM3", n)
df3 <- permutation_test_pairwise(combined_pdui_LV, "rel_count_pdui_adj", "CALM2", "CALM3", n)

permutation_df_LV <- bind_rows(df1, df2, df3) %>%
  mutate(comparison = paste0(gene_a, "_", gene_b))

plot_hist <- function(input_df, selected_comparison) {
  temp_df <- input_df %>%
    filter(comparison == selected_comparison)
  
  gene_a = unique(temp_df$gene_a)
  gene_b = unique(temp_df$gene_b)
  
  temp_df %>%
    ggplot(aes(x = diff)) +
    geom_histogram(color = "black",
                   fill = "white") +
    labs(x = "Median difference",
         y = "Count",
         title = paste0(gene_a, " vs ", gene_b)
         # title = comparison
    )
}

p1 <- plot_hist(permutation_df_LV, "CALM1_CALM2") +
  geom_vline(xintercept = CALM1_CALM2_diff, color = "red")
p2 <- plot_hist(permutation_df_LV, "CALM1_CALM3") +
  geom_vline(xintercept = CALM1_CALM3_diff, color = "red")
p3 <- plot_hist(permutation_df_LV, "CALM2_CALM3") +
  geom_vline(xintercept = CALM2_CALM3_diff, color = "red")

p1 + p2 + p3 + plot_layout(ncol = 1)

range(df1$diff)
range(df2$diff)
range(df3$diff)

df1 %>% filter(abs(diff) >= abs(CALM1_CALM2_diff)) %>% nrow()
df2 %>% filter(abs(diff) >= abs(CALM1_CALM3_diff)) %>% nrow()
df3 %>% filter(abs(diff) >= abs(CALM2_CALM3_diff)) %>% nrow()
```

<br><br>

### LV expression and TE
```{r }
joined_ribo_mrna <- readRDS(here(pdui_te_data, "joined_ribo_mrna.rds"))

cam_te <- joined_ribo_mrna %>% 
  group_by(gene_name) %>% 
  filter(gene_name %in% c("CALM1", "CALM2", "CALM3")) %>% 
  reframe(median_te = median(te_SvH))
rm(joined_ribo_mrna)

# adding TE info
combined_pdui_LV <- combined_pdui_LV %>% 
  left_join(., cam_te, by = c("gene" = "gene_name")) %>% 
  mutate(counts_per_nt_te = counts_per_nt*median_te,
         counts_per_nt_pdui_adj_te = counts_per_nt_pdui_adj*median_te)

# total per sample 
total_counts <- combined_pdui_LV %>% 
  arrange(sample_name) %>% 
  group_by(sample_name) %>% 
  reframe(total_counts_te = sum(counts_per_nt_te),
          total_counts_pdui_adj_te = sum(counts_per_nt_pdui_adj_te))

combined_pdui_LV <- left_join(combined_pdui_LV, total_counts, by = "sample_name") %>% 
  mutate(rel_count_te = counts_per_nt_te/total_counts_te,
         rel_count_pdui_adj_te = counts_per_nt_pdui_adj_te/total_counts_pdui_adj_te)
```

#### Gene expression (PDUI corrected)
```{r Gene expression (PDUI corrected), fig.width = 10, fig.height = 6, out.width = "100%"}
id_order <- combined_pdui_LV %>% 
  filter(gene == "CALM3") %>% 
  arrange(rel_count_pdui_adj) %>% 
  pull(sample_name)

bar_plot_theme <- 
  theme(legend.text = element_text(size = 14,
                                   face = "italic"),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
  axis.text.y = element_text(size = 14),
  axis.title.x = element_text(size = 16),
  axis.title.y = element_text(size = 16), 
  legend.title = element_text(size = 14))
  
p1 <- combined_pdui_LV %>% 
  ggplot(., aes(x = factor(sample_name, levels = id_order), 
                y = rel_count_pdui_adj, fill = gene)) + 
  geom_bar(position="fill", stat="identity") +
  labs(fill = "Gene",
       x = "",
         y = "Expressed per gene", 
       # title = "Gene expression GTEx LV samples (n = 386)"
       ) +
  scale_y_continuous(expand = c(0, 0),
                     breaks = seq(0, 1, by = 0.1),
                     labels = scales::percent_format()) +
  scale_fill_manual(values = selected_colors$gene) +
  bar_plot_theme

combined_pdui_LV %>% 
  group_by(gene) %>% 
  reframe(min = min(rel_count_pdui_adj),
          median = median(rel_count_pdui_adj),
          mean = mean(rel_count_pdui_adj),
          max = max(rel_count_pdui_adj),
          sd = sd(rel_count_pdui_adj)) %>% 
  s_table()
```

```{r Gene expression (PDUI corrected) 8_6, fig.width = 8, fig.height = 6, out.width = "100%", eval = FALSE}
p1
p1 + xlab("Left ventricle samples (n = 386)") 
```
<br><br>

```{r Gene expression (PDUI corrected) 5_4, fig.width = 5, fig.height = 4, out.width = "100%"}
p1 + xlab("Left ventricle samples (n = 386)") 
```


```{r Gene expression (PDUI corrected) 6_6, fig.width = 6, fig.height = 6, out.width = "100%", eval = FALSE}
p1
```
<br><br>


```{r Gene expression (PDUI corrected) 10_4, fig.width = 10, fig.height = 4, out.width = "100%"}
p1
```
<br><br>

```{r Gene expression (PDUI corrected) 10_5, fig.width = 10, fig.height = 5, out.width = "100%", eval = FALSE}
p1
```
<br><br>

#### Translation (PDUI corrected)
```{r Translation (PDUI corrected), fig.width = 10, fig.height = 6, out.width = "100%"}
id_order <- combined_pdui_LV %>% 
  filter(gene == "CALM3") %>% 
  arrange(rel_count_pdui_adj_te) %>% 
  pull(sample_name)

p1 <- combined_pdui_LV %>% 
  ggplot(., aes(x = factor(sample_name, levels = id_order), 
                y = rel_count_pdui_adj_te, fill = gene)) + 
  geom_bar(position="fill", stat="identity") +
  labs(fill = "Gene",
       x = "",
       y = "Translated per gene", 
       # title = "Translation GTEx LV samples (n = 386)"
       ) +
  scale_y_continuous(expand = c(0, 0),
                     breaks = seq(0, 1, by = 0.1),
                     labels = scales::percent_format()) +
  scale_fill_manual(values = selected_colors$gene) +
  bar_plot_theme
p1

combined_pdui_LV %>% 
  group_by(gene) %>% 
  reframe(min = min(rel_count_pdui_adj_te),
          median = median(rel_count_pdui_adj_te),
          mean = mean(rel_count_pdui_adj_te),
          max = max(rel_count_pdui_adj_te),
          sd = sd(rel_count_pdui_adj_te)) %>% 
  s_table()
```
<br><br>

```{r Translation (PDUI corrected) 8_6, fig.width = 8, fig.height = 6, out.width = "100%", eval = FALSE}
p1
```
<br><br>

```{r Translation (PDUI corrected) 6_6, fig.width = 6, fig.height = 6, out.width = "100%"}
p1
```
<br><br>

#### Fraction from each indvidual based on expression and TE
```{r}
combined_pdui_LV %>% 
  group_by(gene) %>% 
  reframe(min = min(rel_count_pdui_adj_te),
          median = median(rel_count_pdui_adj_te),
          mean = mean(rel_count_pdui_adj_te),
          max = max(rel_count_pdui_adj_te),
          sd = sd(rel_count_pdui_adj_te)) %>% 
  s_table()

t1 <- combined_pdui_LV %>% 
  group_by(gene)%>% 
  reframe(median_relative_count = median(rel_count_pdui_adj_te))

CALM1_CALM2_diff <-  t1$median_relative_count[1]-t1$median_relative_count[2]

CALM1_CALM3_diff <- t1$median_relative_count[1]-t1$median_relative_count[3]

CALM2_CALM3_diff <- t1$median_relative_count[2]-t1$median_relative_count[3]
```
<br><br>
  
  * CALM1-CALM2: `r CALM1_CALM2_diff`
  * CALM1-CALM3: `r CALM1_CALM3_diff`
  * CALM2-CALM3: `r CALM2_CALM3_diff`
<br><br>
  
#### Test of signifiance
```{r}
kruskal.test(rel_count_pdui_adj_te ~ gene, data = combined_pdui_LV)
```
<br><br>
  
### All tissue expression
```{r All tissue expression, fig.width = 10, fig.height = 4, out.width = "100%"}
df <- combined_pdui_adj %>% 
  group_by(gene, smtsd) %>% 
  reframe(median = median(rel_count_pdui_adj),
          quant_0.025 = quantile(rel_count_pdui_adj, 0.025),
          quant_0.5 = quantile(rel_count_pdui_adj, 0.5),
          quant_0.975 = quantile(rel_count_pdui_adj, 0.975))

# percentile range 
# data.frame(rel_count_pdui_adj = seq(0, 1000, 1)) %>% 
#   reframe(median = median(rel_count_pdui_adj),
#           quant_0.025 = quantile(rel_count_pdui_adj, 0.025),
#           quant_0.5 = quantile(rel_count_pdui_adj, 0.5),
#           quant_0.975 = quantile(rel_count_pdui_adj, 0.975))


p2 <- df %>% 
  ggplot(., aes(x = smtsd, y = quant_0.5, color = gene)) + 
  geom_point(position = position_dodge(width = 0.8), size = 1.5) +
  geom_linerange(aes(ymin = quant_0.025, ymax = quant_0.975),
                 position = position_dodge(width = 0.8), linewidth = 0.75, alpha = 0.75)

p3 <- p2 + 
  scale_color_manual(values = selected_colors$gene) + 
  scale_y_continuous(expand = c(0, 0),
                     breaks = seq(0, 1, by = 0.2), 
                     labels = scales::percent_format(),
                     limits = c(0,0.85)) +
  theme(
    legend.text = element_text(size = 12, face = "italic"),
    plot.margin = unit(c(0.2, .2, .2, 1.5), "cm"),
    axis.ticks.x = element_blank(),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size = 16),
    axis.title.y = element_text(size = 16), 
    legend.title = element_text(size = 14)) +
  labs(color = "Gene",
       # fill = "Gene",
       x = "", 
       title = "",
       y = "Relative gene \nexpression")

p3
```

```{r All tissue expression 15_10, fig.width = 15, fig.height = 10, out.width = "100%"}
p4 <- p3 + theme(
  axis.text.y = element_text(size = 18),
  axis.text.x = element_text(size = 15),
  axis.title.x = element_text(size = 20),
  axis.title.y = element_text(size = 20), 
  legend.text = element_text(size = 16),
  legend.title = element_text(size = 18))
p4

t1 <- combined_pdui_adj %>% 
  group_by(gene, smtsd) %>%
  reframe(mean_fraction_counts_per_nt_pdui_adj = mean(rel_count_pdui_adj),
          mean_fraction_counts_per_nt = mean(rel_count)) 

write_csv(t1, here(tab_path, "mean_fraction_of_expression_gene_tissue.csv"))

CALM3_mean_sorted <- t1 %>% 
  filter(gene == "CALM3") %>% 
  arrange(mean_fraction_counts_per_nt_pdui_adj) %>% 
  pull(smtsd)
```

```{r All tissue expression 15_7, fig.width = 15, fig.height = 7, out.width = "100%"}
# p4
```

```{r All tissue expression 15_6, fig.width = 15, fig.height = 6, out.width = "100%"}
# p4
```

```{r}
combined_pdui_adj %>% 
  pivot_longer(
    cols = c(rel_count, rel_count_pdui_adj), 
    values_to = "rel_counts_v8", 
    names_to = "measure") %>% 
  group_by(gene, measure) %>%
  reframe(
    median_counts_per_nt = median(rel_counts_v8),
    mean_counts_per_nt = mean(rel_counts_v8)) %>%
  s_table()
```
Summary stats among all `r length(unique(combined_pdui_adj$sampid))` samples are shown above. <br><br>

#### Ratios of expression
```{r}
rel_counts_nested <- combined_pdui_adj %>% 
  dplyr::select(gene, tissue, sampid, rel_count, rel_count_pdui_adj) %>% 
  mutate(ratio = rel_count_pdui_adj/rel_count,
         log2_ratio = log2(ratio)) %>% 
  group_by(gene, tissue) %>% 
  nest

one_sample_wilcox <- function(data) {
  test_result <- wilcox.test(data$log2_ratio, mu = 0, paired = FALSE, alternative = "two.sided")
  broom::tidy(test_result) 
}

one_sample_wilcox_p_values <- rel_counts_nested %>%
  mutate(test_results = map(data, one_sample_wilcox)) %>%
  unnest(test_results) %>%
  dplyr::select(gene, tissue, p.value, statistic, method) %>% 
  ungroup

bonf_threshold <- 0.05/nrow(one_sample_wilcox_p_values)

one_sample_wilcox_p_values <- one_sample_wilcox_p_values %>% 
  mutate(significant = if_else(p.value < bonf_threshold, "sig", "not_sig"))

one_sample_wilcox_p_values %>% 
  dplyr::count(gene, significant) %>% 
  s_table()

tissue_summary_stats <- rel_counts_nested %>% 
  unnest(data) %>%
  group_by(gene, tissue) %>% 
  reframe(min_ratio = min(ratio),
          median_ratio = median(ratio),
          max_ratio = max(ratio),
          min_log2ratio = min(log2_ratio),
          median_log2ratio = median(log2_ratio),
          mean_log2ratio = mean(log2_ratio),
          max_log2ratio = max(log2_ratio)) %>% 
  left_join(., one_sample_wilcox_p_values, by = c("gene", "tissue"))

# tissue_summary_stats %>%
#   filter(abs(median_log2ratio) >= abs(log2(2)) & significant == "sig")

t1 <- tissue_summary_stats %>% 
  group_by(gene) %>% 
  reframe(min_of_medians_log2ratio = min(median_log2ratio),
          median_of_medians_log2ratio = median(median_log2ratio),
          max_of_medians_log2ratio = max(median_log2ratio))
```
<br><br>
  
```{r Ratios of expression, fig.width = 10, fig.height = 10, out.width = "100%"}
p1 <- rel_counts_nested %>% 
  unnest(data) %>% 
  ggplot(aes(x = tissue, y = log2_ratio, color = tissue)) + 
  geom_violin(draw_quantiles = 0.5) + 
  scale_y_continuous(limits = c(-1.75, 1.75)) +
  facet_wrap(~gene, ncol = 1) + 
  theme(legend.position = "none", 
        plot.margin = unit(c(0.2, 0.2, 0.2, .5), "cm"),
        strip.text = element_text(face = "italic", size  = 14),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 14),
        axis.title.y = element_text(size = 16)) +
  scale_x_discrete(labels = function(x) str_replace_all(x, "_", " ")) +
  scale_color_manual(values = selected_colors$tissue) 

df_hlines <- data.frame(gene = cam_genes) %>% 
  left_join(., t1)

p1 + labs(
  x = "", 
  y = expression(
    Log[2](
      frac(
        "Percentage"["PDUI adjusted counts per nt"],
        "Percentage"["Counts per nt"]
        )
      )
    )
  ) +
  geom_hline(data = df_hlines, 
             aes(yintercept = median_of_medians_log2ratio),
             linetype = "dashed")

```
<br><br>

```{r}
t1 %>% s_table()
```


#### Ranked expression
```{r Ranked expression, fig.width = 10, fig.height = 6, out.width = "100%"}
rank_of_median_counts_per_nt_pdui_adj <- combined_pdui_adj %>% 
  dplyr::select(gene, sampid, tissue, rel_count_pdui_adj) %>% 
  group_by(tissue, gene) %>% 
    reframe(median = median(rel_count_pdui_adj)) %>% 
    group_by(tissue) %>% 
    mutate(rank_of_medians = rank(-median)) 

ranks <- rank_of_median_counts_per_nt_pdui_adj %>% 
    ungroup %>% 
    dplyr::count(gene, rank_of_medians) %>%
    pivot_wider(values_from = n, names_from = gene) %>% 
    mutate_all(.funs = ~ifelse(is.na(.), 0, .)) 

ranks %>% 
  s_table()
```
<br><br>

CALM2 is the most expressed gene in all tissues except for two. CALM1 is ranked as the most expressed gene in the two tissue below:
```{r}
CALM1_top_rank <- rank_of_median_counts_per_nt_pdui_adj %>% 
  filter(gene == "CALM1") %>% 
  filter(rank_of_medians == 1) %>% 
  pull(tissue)

rank_of_median_counts_per_nt_pdui_adj %>% 
  filter(tissue %in% CALM1_top_rank) %>% 
  pivot_wider(names_from = tissue, values_from = c(median, rank_of_medians)) %>% 
  dplyr::select(1:3) %>% 
  s_table()
```
<br><br>





### Compositional data analysis
Loading packages
```{r, echo = TRUE}
library(compositions)
library(ICSNP)
```

```{r, eval = FALSE}
writeLines(toBibtex(citation("compositions")), "C:/Users/QJ17ZB/Downloads/composistions.bib")
writeLines(toBibtex(citation("ICSNP")), "C:/Users/QJ17ZB/Downloads/ICSNP.bib")
```


#### PDUI normalised gene expression
```{r, echo = TRUE}
d <- combined_pdui_LV %>% 
  select(person_id = sample_name,
         gene,
         expression = rel_count_pdui_adj)

d_wide <- d |> 
  pivot_wider(id_cols = person_id, 
              names_from = gene, 
              values_from = expression)
V <- ilrBase(D = 3)
d_wide_ilr <- compositions::ilr(d_wide |> select(CALM1, CALM2, CALM3),
                                V = V)

Z <- as.matrix(d_wide_ilr)

# OVERALL:
HotellingsT2(Z, mu = c(0, 0))
```

```{r, echo = TRUE}
# Posthoc:
p.adjust(c(
  t.test(Z[,1], mu=0)$p.value, # CALM1 = CALM2
  t.test(Z[,2], mu=0)$p.value  # CALM3 vs geomean(CALM1, CALM2) 
), method = "BH")

# Posthoc without BH correction
c(
  t.test(Z[,1], mu=0)$p.value, # CALM1 = CALM2
  t.test(Z[,2], mu=0)$p.value  # CALM3 vs geomean(CALM1, CALM2) 
)
```

##### Means 
```{r}
d %>% 
  group_by(gene) %>% 
  reframe(mean_pdui_exp = mean(expression)) %>% 
  s_table()
```



<br><br>
  

#### Translation (PDUI normalised)
```{r, echo = TRUE}
d <- combined_pdui_LV %>% 
  select(person_id = sample_name,
         gene,
         translation = rel_count_pdui_adj_te)

d_wide <- d |> 
  pivot_wider(id_cols = person_id, 
              names_from = gene, 
              values_from = translation)
V <- ilrBase(D = 3)
d_wide_ilr <- compositions::ilr(d_wide |> select(CALM1, CALM2, CALM3),
                                V = V)

Z <- as.matrix(d_wide_ilr)

# OVERALL:
HotellingsT2(Z, mu = c(0, 0))
```

```{r, echo = TRUE}
# Posthoc:
p.adjust(c(
  t.test(Z[,1], mu=0)$p.value, # CALM1 = CALM2
  t.test(Z[,2], mu=0)$p.value  # CALM3 vs geomean(CALM1, CALM2) 
), method = "BH")

# Posthoc without BH correction
c(
  t.test(Z[,1], mu=0)$p.value, # CALM1 = CALM2
  t.test(Z[,2], mu=0)$p.value  # CALM3 vs geomean(CALM1, CALM2) 
)
```

##### Means 
```{r}
d %>% 
  group_by(gene) %>% 
  reframe(mean_pdui_exp = mean(translation)) %>% 
  s_table()
```




------------------------------------------------------------------------
```{r computetime, echo=FALSE}
time <- lubridate::date()
```
Report was finished `r time` <br>


```{r}
sessionInfo()
```


```{r rendering while keeping objects in global environment, echo = FALSE, eval=FALSE}
rmarkdown::render("scripts/expression.Rmd", clean = TRUE, output_dir = "reports/")
```
<br>
