```{r, echo = FALSE}
title <- "Translational efficiency"
```

---
title: `r title`
author: "Steffan Noe Niikanoff Christiansen"
output: html_document
---

### Global options and paths
```{r libraries, echo = FALSE, message=FALSE}
packages <- c("tidyverse", "patchwork", "kableExtra", "here")

invisible(lapply(packages, require, character.only = TRUE)) 
theme_set(theme_bw() + 
            theme(plot.title = element_text(hjust = 0.5)))
```

```{r setup, message=FALSE, echo = FALSE}
# paths
root_path <- here()
projects_path <- str_remove(root_path, "/[A-z]+$")
data_path <- here(projects_path, "public_transcriptomes", "data", "van_heesch_2019_cell")
pdui_te_data <- here(projects_path, "public_transcriptomes", "results", "processed")
fig_path <- here("results", "images", title, "/")

# global options
require("knitr")
opts_knit$set(root.dir = root_path)
opts_chunk$set(warning = TRUE,
               message = FALSE,
               # dpi = 300,
               dpi = 1200,
               # dev = "jpeg",
               dev = "svg",
               fig.width = 6,
               fig.height = 6,
               fig.align="center", 
               fig.path = fig_path,
               echo = FALSE,
               warning = FALSE)

options(scipen = 100, digits = 3, bitmapType = "cairo") # fixes grid.Call warning 
time <- lubridate::date()

# helper functions
s_table <- function (df) {
    kableExtra::kable_styling(knitr::kable(df), bootstrap_options = c("condensed", 
        "striped"), full_width = FALSE, font_size = 14)
}

s_names <- function (df) 
{
    colnames(df) <- tolower(stringr::str_replace_all(colnames(df), 
        c(` \\| ` = "_", ` - ` = "_", ` ` = "_", `|` = "", `-` = "_", 
            `:` = "", `\\(` = "", `\\)` = "", `\\+` = "_", `#` = "")))
    return(df)
}

scroll_table <- function (dataframe) 
{
    kableExtra::scroll_box(kableExtra::kable_styling(knitr::kable(dataframe), 
        "striped", full_width = FALSE, font_size = 14), width = "1000px", 
        height = "500px")
}
```

```{r}
selected_colors <- list(
  gene_name = c(CALM1 = "#1B9E77",
                CALM2 = "#D95F02",
                CALM3 = "#7570B3"))

selected_theme_poster <- theme(
  axis.text.y = element_text(size = 18),
  axis.text.x = element_text(size = 18),
  axis.title.y = element_text(size = 20), 
  axis.title.x = element_text(size = 20), 
  legend.text = element_text(size = 18),
  legend.title = element_text(size = 20),
  title = element_text(size = 18))
```
Rendering of the report was started `r time` <br> Following packages were loaded: `r packages` <br><br>

### Data
Following sheets are available from the `cardiacTranslatome_counts.xlsx` file. The number of transcripts are shown in the table:
```{r Data}
n_transcripts <- function(name) {
  heesch_counts <- readxl::read_xlsx(here(data_path, "cardiacTranslatome_counts.xlsx"), sheet = name)
  temp <- heesch_counts %>% 
    nrow()
  return(temp)
}

n_unique_transcripts <- function(name) {
  heesch_counts <- readxl::read_xlsx(here(data_path, "cardiacTranslatome_counts.xlsx"), sheet = name)
  temp <- heesch_counts %>% 
    distinct(gene_name) %>% 
    nrow()
  return(temp)
}

df <- data.frame(name = c("mRNAseq_CDS_exon_RawCounts", 
                          "Riboseq_CDS_exon_RawCounts", 
                          "mRNAseq_CDS_exon_DESeqNormCount", 
                          "Riboseq_CDS_exon_DESeqNormCoun"))
df <- df %>% 
  mutate(transcripts = map(name, n_transcripts),
         unique_transcripts = map(name, n_unique_transcripts))
df %>% 
  s_table()
```

The **Differential expression analysis** section of the original paper (10.1016/j.cell.2019.05.010) states following: *For this analysis we include all genes that we consider to be translated, i.e., matching the criteria described in the ‘detecting active translation’ section (n = 11,387)*

**Detecting actively translated ORFs:** *The ﬁnal list of translation events (including all uORFs, downstream ORFs (dORFs) and lncRNA sORFs) was stringently ﬁltered requiring the translated gene to have an average RNA RPKM  1 and be detected as translated in at least 10/80 human, 3/10 rat and 2/6 mouse heart samples. Additionally, each speciﬁc ORF was required to have an identical translation termination codon in at least 5/81 human, 2/10 rat and 2/6 mouse heart samples.*
<br><br>

### Exploration of sample data
```{r Exploration of sample data, fig.width = 10, fig.height = 6, out.width = "100%"}
van_h_s1 <- readxl::read_xlsx(here(data_path, "mmc1.xlsx"), sheet = 1, n_max = 80) %>% s_names
colnames(van_h_s1)[2] <- "general_sample_id"

van_h_s1 <- van_h_s1 %>% 
  mutate(age_at_surgical_procedure = as.numeric(age_at_surgical_procedure))
```

##### Sex
```{r}
van_h_s1 %>% 
  dplyr::count(xist_sex, `reported_sex_m_/_f`) %>% 
  s_table
```
<br><br>

##### Age
```{r Age, fig.width = 10, fig.height = 6, out.width = "100%"}
van_h_s1 %>% 
  na.omit(age_at_surgical_procedure) %>% 
  reframe(min = min(age_at_surgical_procedure),
          median = median(age_at_surgical_procedure),
          max = max(age_at_surgical_procedure)) %>% 
  s_table
```
There were no age information on three individuals. The sum stats for the remaining individuals are shown above. Median age is reported as 43.6 years in original paper.
<br><br>

##### Disease and circumstances for tissue collection
```{r}
abbr <- readxl::read_xlsx(here(data_path, "mmc1.xlsx"), sheet = 1, n_max = 1) %>% 
  select(7) %>% 
  names()

tab <- van_h_s1 %>% 
  dplyr::count(`disease_status_dcm_/_unaffected_control`, 
               `sample_obtained_during_surgical_procedure_htx_=_heart_transplantation;_lvad_=_left_ventricular_assistant_device_implantation;_vsarr_=_valve_sparing_aortic_root_replacement_`) 

colnames(tab) <- c("phenotype", "tissue collection")
  
tab %>% 
  s_table
```
`r abbr` <br>


### Detection of open reading frames
```{r Detection of open reading frames, fig.width = 10, fig.height = 6, out.width = "100%"}
orfs <- readxl::read_xlsx(here(data_path, "mmc1.xlsx"), sheet = 2, skip = 1) %>% s_names
cam_genes <- c("CALM1", "CALM2", "CALM3")

t1 <- orfs %>% 
  filter(gene_symbol %in% cam_genes) %>% 
  mutate(aa = nchar(`orf_peptide_sequence_aa;_max_32,000_characters`)) %>% 
  select(gene_symbol, 
         transcript_id,
         aa,
         "_of_samples_with_the_exact_same_orf", 
         "_of_samples_with_the_same_orf_stop",
         "_of_samples_with_the_same_gene_translated",
         orf_p_value) %>% 
  mutate(orf_p_value = format(orf_p_value, digits = 3, scientific = 1)) %>% 
  arrange(gene_symbol)
colnames(t1) <- c("Gene name", "Transcript", "n AA", "n exact same ORF", "n exact same ORF stop", "n same gene translated", "ORF p-value")

t1 %>% 
  s_table
```
Notice: <br> 
  
  * There are only one reading frame for each of the CALM genes
<br>

The distribution of all open reading frames are shown below:
```{r Distribution of open reading frames, fig.width = 10, fig.height = 6, out.width = "100%"}
t2 <- orfs %>% 
  dplyr::count(gene_symbol, name = "n_transcripts", sort = TRUE)

t3 <- t2 %>% 
  dplyr::count(n_transcripts, name = "n_occurences")

t3 %>% 
  ggplot(aes(x = n_transcripts, y = n_occurences)) +
  geom_bar(stat = "identity") +
  scale_y_continuous()

t2 %>%
  slice_max(n_transcripts, n = 10) %>%
  s_table()
```
Top 10 genes with most transcripts are shown above.
<br><br>

### Exploration of raw and normalised counts
```{r loading counts}
return_counts <- function(name) {
  heesch_counts <- readxl::read_xlsx(here(data_path, "cardiacTranslatome_counts.xlsx"), sheet = name) %>% 
    s_names
  subset <- heesch_counts %>% 
    filter(gene_biotype == "protein_coding") %>% 
    pivot_longer(cols = 4:ncol(.), 
                 names_to = "sample_name", 
                 values_to = "counts") %>% 
    mutate(sheet = name)
  return(subset)
}

counts_df <- bind_rows(
  return_counts("mRNAseq_CDS_exon_RawCounts") %>% 
    mutate(method = "mRNAseq",
           status = "raw"),
  return_counts("Riboseq_CDS_exon_RawCounts") %>% 
    mutate(method = "riboseq",
           status = "raw"),
  return_counts("mRNAseq_CDS_exon_DESeqNormCount") %>% 
    mutate(method = "mRNAseq",
           status = "normalised"),
  return_counts("Riboseq_CDS_exon_DESeqNormCoun") %>% 
    mutate(method = "riboseq",
           status = "normalised")) %>% 
  mutate(status = factor(status, levels = c("raw", "normalised")),
         id = str_extract(sample_name, "[0-9]{3}"))
```

```{r Exploration of counts from protein coding genes, fig.width = 10, fig.height = 5, out.width = "80%", warnings = TRUE}
# range(counts_df$counts)
axis_limits <- c(0, 175000)

counts_df %>% 
  dplyr::select(-sheet, gene_biotype) %>%
  pivot_wider(values_from = counts, names_from = status) %>% 
  na.omit() %>% 
  ggplot(aes(x = raw, y = normalised)) + 
  geom_bin2d(bins = 250) + 
  geom_abline(intercept = 0, slope = 1, color = "red", linetype= "dashed", linewidth = 0.75) + 
  scale_y_continuous(limits = axis_limits) +
  scale_x_continuous(limits = axis_limits) +
  scale_fill_gradient(trans = "log10") + 
  facet_wrap(~ method) +
  coord_fixed() +
  labs(x = "Raw counts",
       y = "Normalised counts",
       fill = "Number of occurences")

counts_df %>% 
  dplyr::select(-sheet, gene_biotype) %>%
  group_by(gene_name, method, status) %>%
  dplyr::slice(seq(9, 79, by = 10)) %>%  
  pivot_wider(values_from = counts, names_from = status) %>% 
  na.omit() %>% 
  ggplot(aes(x = raw, y = normalised, color = id)) + 
  geom_point(size = 0.2) + 
  geom_abline(intercept = 0, slope = 1, color = "black", linetype= "dashed", linewidth = 0.75) + 
  scale_y_continuous(limits = axis_limits) +
  scale_x_continuous(limits = axis_limits) +
  facet_wrap(~ method) +
  coord_fixed() +
  labs(x = "Raw counts",
       y = "Normalised counts")
```

##### Sum of counts
```{r Sum of counts, fig.width = 8, fig.height = 5, out.width = "80%", warnings = TRUE}
sum_of_counts <- counts_df %>% 
  group_by(method, status, id) %>% 
  reframe(sum_of_counts = sum(counts)) 

sum_of_counts %>% 
  ggplot(aes(y = sum_of_counts, x = status, color = status)) +
  geom_boxplot(outlier.shape = NA) + 
  facet_wrap(~method, ncol = 2, scales = "free_x") +
  geom_jitter(width = 0.05, size = 0.5, alpha = 1, height = 0)
```
<br>

```{r}
sum_of_counts %>% 
  group_by(method, status) %>% 
  reframe(median_sum_of_counts = median(sum_of_counts)) %>% 
   mutate(across(where(is.numeric), ~ format(., big.mark=","))) %>% 
  s_table()
```
<br><br>
  
#### Mitochondrial counts
```{r Mitochondrial counts, fig.width = 10, fig.height = 6, out.width = "100%"}
counts_df %>% 
  filter(str_detect(gene_name, "^MT-")) %>% 
  dplyr::count(method, status) %>% 
  s_table()

counts_df %>% 
  filter(str_detect(gene_name, "^MT-")) %>%
  group_by(id) %>% 
  reframe(sum_mt_counts = sum(counts)) %>% 
  slice_max(sum_mt_counts, n = 5) %>% 
  s_table()
```
Mitochondrial genes (`MT-` prefix) is only present in raw data and not in normalised data. Top five are shown above
<br><br>
  


### Processing of counts for analysis of differential expression

According to the Van Heesch paper - **Differential expression analysis**, following steps were conducted before DE analysis:

*To allow for proper comparison and integration of mRNA-seq and Ribo-seq data, all mRNA-seq quantiﬁcations were derived from CDS-mapped, single-end reads trimmed from 101nt to ribosomal footprint sizes. Expression quantiﬁcation was followed by read normalization, size factor estimation (on mRNA-seq and & Ribo-seq simultaneously) and differential expression analysis between DCM and unaffected control samples using DESeq2 v1.12.4 (Love et al., 2014).*


#### Estimation of size factor
As suggested by SvH and evident from [Witte et al (2021), Genome Biology](https://github.com/jorruior/witte_et_al_2021/blob/main/todo/Preprocessing/2_Make_TE_Matrix_normalized.R), it is necessary to estimate size factor from RNA seq and Riboseq at the same time! The resulting size factors per sample and analysis are shown below:

```{r Estimation of size factor, fig.width = 10, fig.height = 6, out.width = "70%"}
library(DESeq2)
read_from_xlsx <- function(sheet){
  data <-  readxl::read_xlsx(
    here(data_path, "cardiacTranslatome_counts.xlsx"), 
    sheet = sheet)
  countdata <- as.matrix(data[, -c(1:3)])
  
  rownames(countdata) <- pull(data[, 1])
  return(countdata)
}

mRNA_raw <- read_from_xlsx("mRNAseq_CDS_exon_RawCounts")
ribo_raw <- read_from_xlsx("Riboseq_CDS_exon_RawCounts")

pool_sizeFactor <- estimateSizeFactorsForMatrix(cbind(mRNA_raw, ribo_raw))

pool_size_df <- data.frame(pool_sizeFactor) %>% 
  rownames_to_column(., "sample_name") %>% 
  as_tibble() %>% 
  separate(col = "sample_name", 
           into = c("species", "tissue", "id", "status", "data_type"), sep="_", remove = FALSE)

pool_size_df %>% 
  ggplot(aes(x = data_type, y = pool_sizeFactor, group = id)) + 
  geom_point() + 
  geom_line()
```

Notice that the scaling factor for ribo are lower than for mRNA in all individuals except for three.

```{r , fig.width = 10, fig.height = 6, out.width = "70%"}
pool_size_df %>% 
  select(id, data_type, pool_sizeFactor) %>% 
  pivot_wider(names_from = data_type, values_from = pool_sizeFactor) %>% 
  filter(Ri > mR) %>% 
  s_table()

# data.frame(mRNA_raw = colSums(mRNA_raw),
#            ribo_raw = colSums(ribo_raw)) %>% 
#   filter(mRNA_raw < ribo_raw)
```

This difference is not exclusively due to size of the library because the sample with id 30 is the only sample with higher raw Ribo count compared to mRNA count.

```{r , fig.width = 10, fig.height = 6, out.width = "70%"}
pool_size_wide_df <- pool_size_df %>% 
  select(id, data_type, pool_sizeFactor) %>% 
  pivot_wider(names_from = data_type, values_from = pool_sizeFactor)
```

Among the data from the 80 samples, the size factor for mRNA and ribo seq were \< 1 for `r pool_size_wide_df %>% filter(mR < 1) %>% nrow()` and `r pool_size_wide_df %>%  filter(Ri < 1) %>% nrow()` samples, respectively.

```{r}
pool_size_wide_df %>% 
  filter(mR < 1) %>% 
  s_table()
```

<br><br>

##### Reproducing normalisation results 
The raw counts are normalised with following approach: <br>
`x/pool_sizeFactor[1:ncol(mRNA_raw)]`.
```{r}
mRNA_norm_snnc <- t(apply(mRNA_raw, 1, function(x) {x/pool_sizeFactor[1:ncol(mRNA_raw)]}))
ribo_norm_snnc <- t(apply(ribo_raw, 1, function(x) {x/pool_sizeFactor[81:160]}))

mRNA_norm <- read_from_xlsx("mRNAseq_CDS_exon_DESeqNormCount")
ribo_norm <- read_from_xlsx("Riboseq_CDS_exon_DESeqNormCoun")

norm_data_genes <- intersect(rownames(mRNA_norm), rownames(ribo_norm))
```

Normalised by SNNC:
```{r}
mRNA_norm_snnc[1:4,1:5] %>% 
  s_table()
```
<br>

Uploaded normalized data from paper:
```{r}
mRNA_norm[1:4,1:5] %>% 
  s_table()
```
Notice that there the small differences in the decimal numbers. Possibly caused by different number of included genes in the object normalised by SvH (`r nrow(mRNA_norm)`) and by SNNC (`r nrow(mRNA_norm_snnc)`).<br>

```{r }
count_to_tidy_df <- function(count_matrix) {
  count_matrix %>% 
    data.frame() %>% 
    rownames_to_column(., "gene_id") %>% 
    pivot_longer(2:ncol(.), names_to = "sample_name", values_to = "counts")
}

mRNA_norm_joined <- inner_join(count_to_tidy_df(mRNA_norm),
                               count_to_tidy_df(mRNA_norm_snnc),
                               by = c("gene_id", "sample_name"),
                               suffix = c("_norm_SvH", "_norm_SNC"))

ribo_norm_joined <- inner_join(count_to_tidy_df(ribo_norm),
                               count_to_tidy_df(ribo_norm_snnc),
                               by = c("gene_id", "sample_name"),
                               suffix = c("_norm_SvH", "_norm_SNC"))
```

```{r Comparing normalisations, fig.width = 10, fig.height = 5, out.width = "100%", warnings = TRUE}
axis_limits <- c(0, 10^6)
plot_comparison <- function(joined_data) {
  joined_data %>% 
    ggplot(aes(x = counts_norm_SvH, y = counts_norm_SNC)) + 
    geom_bin2d(bins = 250) + 
    geom_abline(intercept = 0, slope = 1, color = "grey", linetype= "dashed", linewidth = 0.75) + 
    scale_y_continuous(limits = axis_limits) +
    scale_x_continuous(limits = axis_limits) +
    scale_fill_gradient(trans = "log10") + 
    coord_fixed()
}

plot_comparison(mRNA_norm_joined) + ggtitle("Counts among 11,387 genes (mRNA)") +
  plot_comparison(ribo_norm_joined) + ggtitle("Counts among 11,387 genes (ribo)")
```
The plot are based on `r length(unique(ribo_norm_joined$gene_id))` transcripts among `r length(unique(ribo_norm_joined$sample_name))` samples.
<br><br>

Notice: <br> 
  
  * Since the DeSeq method **do not** take length into account we would expect that longer genes have higher count values. Since only CDS mapped reads were included in analysis, there should be no difference between CALM1-3. I.e. the normalised counts should be able to be used for further analyses. 
<br><br>
  
```{r , fig.height=5, fig.width=10}
gene_names <- return_counts("mRNAseq_CDS_exon_DESeqNormCount") %>% 
  select(gene_id:gene_biotype) %>% 
  distinct()

joined_df <- full_join(
  ribo_norm_joined %>% 
    separate(col = "sample_name", 
             into = c("species", "tissue", "id", "status", "data_type"), sep="_", remove = TRUE) %>% 
    select(-data_type),
  mRNA_norm_joined %>% 
    separate(col = "sample_name", 
             into = c("species", "tissue", "id", "status", "data_type"), sep="_", remove = TRUE) %>% 
    select(-data_type),
  by = c("gene_id", "species", "tissue", "id", "status"),
  suffix = c("_ribo", "_mRNA")
) %>% 
  mutate(te_SvH = counts_norm_SvH_ribo/counts_norm_SvH_mRNA)

joined_df <- left_join(joined_df, gene_names, by = "gene_id")
```

```{r, eval = FALSE}
saveRDS(joined_df, here("results", "processed", "joined_ribo_mrna.rds"))
```

```{r}
joined_df <- readRDS(here(pdui_te_data, "joined_ribo_mrna.rds"))

joined_df <- joined_df %>% 
  left_join(., van_h_s1 %>% 
              mutate(id = str_extract(general_sample_id, "[0-9]{3}")) %>% 
              select(id, xist_sex, age_at_surgical_procedure))

```


### Translational efficiency
*Translational efficiency (TE) estimations were calculated by taking the ratio of Ribo-seq over mRNA-seq counts. This approach yields independent TEs for each individual sample and gene* (**Translational efficiency estimation** section in Van Heesch 2019.). Similar analyses are available from [Chothani er al 2019](https://currentprotocols.onlinelibrary.wiley.com/doi/10.1002/cpmb.108)
```{r TE per individual, fig.width = 10, fig.height = 6, out.width = "100%", warning = TRUE}
values <- c(1, 2, 4, 8, 16, 32)
y_axis <- c(1/values^2, 0, values^2)
plot_violin <- function(te_df) {
  te_df %>% 
    drop_na() %>% 
    group_by(id) %>% 
    ggplot(aes(x = id, y = te_SvH)) +
    geom_violin(draw_quantiles = c(0.5)) +
    scale_y_continuous(trans = "log2", labels = y_axis, breaks = y_axis) +
    theme(axis.text.x = element_text(angle = 45,
                                     margin = margin(4, 0, 0, 0),
                                     hjust = 1)) +
    geom_hline(yintercept = 1, color = "red", linetype = "dashed")
}

p1_samp <- joined_df %>% distinct(id) %>% dplyr::slice(1:40) %>% pull

p1 <- joined_df %>% 
  filter(id %in% p1_samp) %>% 
  plot_violin + labs(title = "Translational efficiency per individual", x = "")

p2 <- joined_df %>% 
  filter(!id %in% p1_samp) %>% 
  plot_violin

p1 + p2 + plot_layout(ncol = 1)

median_TE_per_id <- joined_df %>%
  drop_na() %>% 
  group_by(id) %>% 
  reframe(median_TE_per_ind = median(te_SvH))

median_TE_ungrouped <- joined_df %>% 
  na.omit() %>% 
  filter(te_SvH != Inf) %>% 
  reframe(median_te = median(te_SvH)) %>% 
  pull()
```
The median TE per individual ranges from `r min(median_TE_per_id$median_TE_per_ind)` to `r max(median_TE_per_id$median_TE_per_ind)` (median of medians: `r median(median_TE_per_id$median_TE_per_ind)`). Median of all genes (not grouped by ID): `r median_TE_ungrouped` The dashed line corresponds to a TE of 1. <br><br>


##### TE among three main troponins (Svh approach)
```{r TE among three main troponins (Svh approach), fig.width = 5, fig.height = 5, out.width = "60%"}
joined_df %>% 
  filter(gene_name %in% c("TNNC1", "TNNI3", "TNNT2")) %>% 
  ggplot(aes(x = gene_name, y = te_SvH, color = gene_name)) +
  # geom_boxplot(outlier.shape = NA) + 
  geom_violin() +
  geom_jitter(width = 0.25, size = 1, alpha = 1, height = 0) +
  theme(axis.text.x = element_text(face = "italic")) +
  scale_y_continuous(breaks = seq(1, 3, by = 1)) +
  scale_color_manual(values = c("brown", "orange", "red")) +
  labs(y = "Translational efficiency", 
       x = "",
       color = "Gene name")
```
Notice resemblance with [Fig S2G](https://www.sciencedirect.com/science/article/pii/S0092867419305082#figs2) in van Heesch paper. 
<br><br>

##### TE among CALM1-3 
```{r TE among CALM1-3 5_5, fig.width = 5, fig.height = 5, out.width = "60%"}
p1 <- joined_df %>% 
  filter(gene_name %in% c("CALM1", "CALM2", "CALM3")) %>% 
  mutate(xist_sex = str_replace_all(xist_sex, c("f$" = "Female",
                                                "m$" = "Male")),
         status = str_replace_all(status, c("C" = "Control",
                                            "D" = "Dilated cardiomyopathy"))) %>% 
  ggplot(aes(x = gene_name, y = te_SvH, color = gene_name)) +
  geom_violin(draw_quantiles = 0.5) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.7, height = 0) +
  theme(legend.text = element_text(face = "italic"),
        axis.text.x = element_text(face = "italic", size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.y = element_text(size = 16),
        legend.position = "none") +
  scale_y_continuous(breaks = seq(1, 7, by = 2)) +
  scale_color_manual(values = selected_colors$gene_name) +
  labs(y = "Translational efficiency", 
       x = "",
       color = "Gene name")
p1
```
The rank of medians are unchanged if analysis is stratified by disease status. <br>

```{r TE among CALM1-3 5_6, fig.width = 5, fig.height = 6, out.width = "60%", eval = FALSE}
p1 + selected_theme_poster
```
<br><br>

```{r TE among CALM1-3 5_4, fig.width = 5, fig.height = 4, out.width = "60%"}
p1
```
<br><br>



#### TE among CALM1-3 (sex and status)
```{r TE among CALM1-3 (sex and status) 10_5, fig.width = 10, fig.height = 5, out.width = "60%"}
p1 + facet_wrap(~status, ncol = 2) + theme(strip.text = element_text(size = 14))
p1 + facet_wrap(~xist_sex, ncol = 2) + theme(strip.text = element_text(size = 14))


p1 + facet_wrap(~xist_sex + status, ncol = 2)
joined_df %>% 
  distinct(id, xist_sex, status) %>%
  dplyr::count(xist_sex, status)
```

Median age per group is shown below:
```{r}
joined_df %>% 
  distinct(id, xist_sex, status, age_at_surgical_procedure) %>% 
  drop_na(age_at_surgical_procedure) %>% 
  group_by(xist_sex) %>%
  reframe(median_age = median(age_at_surgical_procedure)) %>% 
  s_table

joined_df %>% 
  distinct(id, xist_sex, status, age_at_surgical_procedure) %>% 
  drop_na(age_at_surgical_procedure) %>% 
  group_by(status) %>%
  reframe(median_age = median(age_at_surgical_procedure)) %>% 
  s_table
```
<br><br>
  
###### Sum stats of TE
```{r}
median_TE_per_gene <- joined_df %>%
  drop_na() %>% 
  group_by(gene_name) %>% 
  reframe(median_TE_per_gene = median(te_SvH)) %>% 
  mutate(rank_of_medians = rank(-median_TE_per_gene),
         percentile_rank_of_medians = percent_rank(rank_of_medians))

joined_df_CALM <- joined_df %>% 
  filter(gene_name %in% c("CALM1", "CALM2", "CALM3"))

t1 <- joined_df_CALM %>%
  group_by(gene_name) %>% 
  reframe(mean_te = mean(te_SvH),
          sd_te = sd(te_SvH)) %>% 
  left_join(., median_TE_per_gene)

t1 %>% 
  s_table()
```

Notice: <br> 
  
  * TE are relatively high for all three genes. 
  * High variability among TE
  * Increase in median values for CALM1 compared to CALM2: `r (t1$median_TE_per_gene[1]-t1$median_TE_per_gene[2])/t1$median_TE_per_gene[2]*100` %
  * Increase in median values for CALM1 compared to CALM3: `r (t1$median_TE_per_gene[1]-t1$median_TE_per_gene[3])/t1$median_TE_per_gene[3]*100` %
  * Increase in median values for CALM2 compared to CALM3: `r (t1$median_TE_per_gene[2]-t1$median_TE_per_gene[3])/t1$median_TE_per_gene[3]*100` %
<br>

  
#### Test of difference between the groups
Notice: <br> 
  
  * Fractions are assumed not to be normally distributed
  * Samples are paired
<br><br>

##### Paired comparison (Wilcoxon signed rank test)
```{r}
joined_df_CALM_wide <- joined_df_CALM %>% 
  select(id, xist_sex, status, gene_name, te_SvH) %>%
  pivot_wider(names_from = gene_name, values_from = te_SvH)
```
P-values are not adjusted. Bonferroni corrected significance threshold is 0.05/3 = 0.017 <br><br>

```{r}
options(digits = 6) # fixes grid.Call warning 
wilcox.test(joined_df_CALM_wide$CALM1, joined_df_CALM_wide$CALM2, paired = TRUE)
wilcox.test(joined_df_CALM_wide$CALM1, joined_df_CALM_wide$CALM3, paired = TRUE)
wilcox.test(joined_df_CALM_wide$CALM2, joined_df_CALM_wide$CALM3, paired = TRUE)
```

###### Wilcoxon of subgroups
ctl
```{r}
joined_df_CALM_wide_ctl <- joined_df_CALM_wide %>% 
  filter(status == "C")

wilcox.test(joined_df_CALM_wide_ctl$CALM1, joined_df_CALM_wide_ctl$CALM2, paired = TRUE)
wilcox.test(joined_df_CALM_wide_ctl$CALM1, joined_df_CALM_wide_ctl$CALM3, paired = TRUE)
wilcox.test(joined_df_CALM_wide_ctl$CALM2, joined_df_CALM_wide_ctl$CALM3, paired = TRUE)
```
<br>
  
DCM
```{r}
joined_df_CALM_wide_dcm <- joined_df_CALM_wide %>% 
  filter(status == "D")

wilcox.test(joined_df_CALM_wide_dcm$CALM1, joined_df_CALM_wide_dcm$CALM2, paired = TRUE)
wilcox.test(joined_df_CALM_wide_dcm$CALM1, joined_df_CALM_wide_dcm$CALM3, paired = TRUE)
wilcox.test(joined_df_CALM_wide_dcm$CALM2, joined_df_CALM_wide_dcm$CALM3, paired = TRUE)
```
<br>

Female
```{r}
joined_df_CALM_wide_f <- joined_df_CALM_wide %>% 
  filter(xist_sex == "f")

wilcox.test(joined_df_CALM_wide_f$CALM1, joined_df_CALM_wide_f$CALM2, paired = TRUE)
wilcox.test(joined_df_CALM_wide_f$CALM1, joined_df_CALM_wide_f$CALM3, paired = TRUE)
wilcox.test(joined_df_CALM_wide_f$CALM2, joined_df_CALM_wide_f$CALM3, paired = TRUE)
```
<br>

Male
```{r}
joined_df_CALM_wide_m <- joined_df_CALM_wide %>% 
  filter(xist_sex == "m")

wilcox.test(joined_df_CALM_wide_m$CALM1, joined_df_CALM_wide_m$CALM2, paired = TRUE)
wilcox.test(joined_df_CALM_wide_m$CALM1, joined_df_CALM_wide_m$CALM3, paired = TRUE)
wilcox.test(joined_df_CALM_wide_m$CALM2, joined_df_CALM_wide_m$CALM3, paired = TRUE)
```

<br><br>
  
##### Correlations of TE among CALM1-3
```{r Correlations of TE among CALM1-3, fig.width = 10, fig.height = 5, out.width = "100%", eval = TRUE}
te_wide <- joined_df_CALM %>% 
  select(gene_name, id, te_SvH, status) %>% 
  pivot_wider(names_from = gene_name, values_from = te_SvH)  
  
plot_te <- function(gene_a, gene_b) {
  
  # cor
  correlation <- broom::tidy(
    cor.test(
      te_wide %>% pull(gene_a), 
      te_wide %>% pull(gene_b), 
      method = "spearman"))
  # print(correlation)
  
  # plotting
  te_wide %>%
    ggplot(aes(x = get(paste(gene_a)),
               y = get(paste(gene_b)),
               # color = status
    )
    ) +
    geom_point(size = 1, alpha = 0.5) +
    theme(axis.text.x = element_text(size = 15),
          axis.text.y = element_text(size = 15),
          axis.title.x = element_text(size = 16, face = "italic"),
          axis.title.y = element_text(size = 16, face = "italic")) +
    coord_fixed() +
    scale_y_continuous(limits = selected_limits, breaks = seq(1, 7, 2)) +
    scale_x_continuous(limits = selected_limits, breaks = seq(1, 7, 2)) +
    labs(y = gene_b,
         x = gene_a) +
    annotate("text",
             label = paste(
               "rho == ", round((correlation$estimate), digits = 2)),
             x = 0.70*selected_limits[2],
             y = 0.95*selected_limits[2],
             hjust = 0,
             vjust = 1, 
             parse = TRUE, 
             size = 5) +
    annotate("text",
             label = paste(
               "italic(p) == ", round((correlation$p.value), digits = 3)),
             x = 0.70*selected_limits[2],
             y = 0.85*selected_limits[2],
             hjust = 0,
             vjust = 1, 
             parse = TRUE, 
             size = 5)
}

selected_limits = c(0.5, 7.25)
p1 <- plot_te("CALM1", "CALM2")
p2 <- plot_te("CALM1", "CALM3")
p3 <- plot_te("CALM2", "CALM3")

# p1 + p2 + p3 + plot_layout(ncol = 3, guides = "collect")
```


```{r Correlations of TE among CALM1-3 2, fig.width = 5, fig.height = 10, out.width = "60%", eval = TRUE}
p1 + p2 + p3 + plot_layout(ncol = 1, guides = "collect")
```


```{r}
df <- data.frame(gene_a = c("CALM1", "CALM1", "CALM2"),
                 gene_b = c("CALM2", "CALM3", "CALM3"))

return_cor_p <- function(gene_a, gene_b) {
  cor.test(te_wide[[gene_a]], te_wide[[gene_b]], method = cor_method)$p.value
}

return_cor_est <- function(gene_a, gene_b) {
  cor.test(te_wide[[gene_a]], te_wide[[gene_b]], method = cor_method)$estimate
}
```
Corresponding correlations are shown in the tables below. Notice that Bonferroni corrected significance threshold is `0.05/3` = `r 0.05/3`. 
<br> 

###### Pearson correlations
```{r Pearson correlations, fig.width = 10, fig.height = 5, out.width = "100%"}
cor_method = "pearson"
df %>% 
  mutate(p_value = map2_dbl(gene_a, gene_b, return_cor_p),
         estimate = map2_dbl(gene_a, gene_b, return_cor_est)) %>% 
  s_table()
```
<br><br>

###### Spearman correlations
```{r Spearman correlations, fig.width = 10, fig.height = 5, out.width = "100%"}
cor_method = "spearman"
df %>% 
  mutate(p_value = map2_dbl(gene_a, gene_b, return_cor_p),
         estimate = map2_dbl(gene_a, gene_b, return_cor_est)) %>% 
  s_table()
```
<br><br>

#### TE heatmaps
Heatmap with absolute values and with z-scores to the left
```{r TE heatmaps, fig.width = 10, fig.height = 6, out.width = "80%"}
te_matrix <- joined_df %>%
  filter(gene_name %in% cam_genes) %>%
  dplyr::select(gene_name, id, te_SvH) %>%
  pivot_wider(names_from = gene_name, values_from = te_SvH) %>%
  dplyr::select(cam_genes) %>%
  data.matrix()
colnames(te_matrix) <- cam_genes

cal_z_score <- function(x){
  (x - mean(x)) / sd(x)
}

# te_matrix[1:3, ]
# t(apply(te_matrix[1:3,], 1, scale))
# t(apply(te_matrix[1:3,], 1, cal_z_score)) # row based standardisation (identical with scale)
# apply(te_matrix[1:3, ], 2, cal_z_score) # col based standardisation

# Column wide standardisation (compare the two outputs above)
te_matrix_z <- apply(te_matrix, 2, cal_z_score)

p1 <- pheatmap::pheatmap(te_matrix, silent = TRUE)
p2 <- pheatmap::pheatmap(te_matrix_z, silent = TRUE)
gridExtra::grid.arrange(p1$gtable, p2$gtable, ncol = 2)
```
<br><br>

------------------------------------------------------------------------

```{r computetime, echo=FALSE}
time <- lubridate::date()
```

Report was knitted `r time` <br>

#### Session info
```{r}
sessionInfo()
```

```{r rendering while keeping objects in global environment, echo = FALSE, eval=FALSE}
rmarkdown::render("scripts/translational_efficiency.rmd", clean = TRUE, output_dir = "reports/")
```

<br>
