```{r, echo = FALSE}
title <- "variants_and_clinical_data"
```

---
title: `r title`
author: "Steffan Noe Niikanoff Christiansen"
output: html_document
---


### Global options and paths
```{r Global options and paths, echo = FALSE, message=FALSE}
# libraries
packages <- c("tidyverse", "patchwork", "kableExtra", "here", "knitr")
invisible(lapply(packages, require, character.only = TRUE)) 

# plotting themes
invisible(lapply(packages, require, character.only = TRUE)) 

selected_theme <- theme(axis.text.y = element_text(size = 14),
      axis.text.x = element_text(size = 14),
      axis.title.y = element_text(size = 16), 
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 14),
      strip.text = element_text(size = 12))

theme_set(theme_bw() + 
            theme(strip.background = element_rect(fill="white"),
                  plot.title = element_text(hjust = 0.5)))

# paths
root_path <- here()
projects_path <- str_remove(root_path, "/[A-z]+$")
gnomAD_path <- here(projects_path, "public_wes_wgs", "data", "constraint", "gnomad_v4")
title <- str_replace_all(title, " ", "_")
fig_path <- here("results", "images", title, "/")
tab_path <- here("results", "tables", title, "/")

# global options
opts_knit$set(root.dir = root_path)
opts_chunk$set(message = FALSE,
               # dpi = 300,
               dpi = 1200,
               # dev = "jpeg",
               dev = "svg",
               fig.width = 6,
               fig.height = 6,
               fig.align="center", 
               fig.path = fig_path,
               echo = FALSE,
               warning = FALSE)

options(scipen = 100, digits = 3, bitmapType = "cairo") 

selected_colors <- list(
  gene = c(CALM1 = "#1B9E77",
           CALM2 = "#D95F02",
           CALM3 = "#7570B3"))

time <- lubridate::date()

# helper functions
s_table <- function (df) {
    kableExtra::kable_styling(knitr::kable(df), bootstrap_options = c("condensed", 
        "striped"), full_width = FALSE, font_size = 14)
}

s_names <- function (df) 
{
    colnames(df) <- tolower(stringr::str_replace_all(colnames(df), 
        c(` \\| ` = "_", ` - ` = "_", ` ` = "_", `|` = "", `-` = "_", 
            `:` = "", `\\(` = "", `\\)` = "", `\\+` = "_", `#` = "")))
    return(df)
}

scroll_table <- function (dataframe) 
{
    kableExtra::scroll_box(kableExtra::kable_styling(knitr::kable(dataframe), 
        "striped", full_width = FALSE, font_size = 14), width = "1000px", 
        height = "500px")
}
```

Rendering of the report was started `r time` <br> Following packages
were loaded: `r packages` <br><br>


### Clinical characteristics of missense carriers
```{r Clinical characteristics of missense carriers, fig.width = 6, fig.height = 6, out.width = "100%", eval = TRUE}
cam_genes <- c("CALM1", "CALM2", "CALM3")
clin_char_df <- data.frame(gene = cam_genes,
                           n_carriers = c(52, 53, 35),
                           cardiac_events = c(46, 37, 20),
                           MAE = c(34, 30, 13)) %>% 
  mutate(gene = factor(gene),
         frac_CE = cardiac_events/n_carriers,
         frac_MAE = MAE/n_carriers,
         free_from_CE = n_carriers - cardiac_events,
         free_from_MAE = n_carriers - MAE)

clin_char_long <- clin_char_df %>%
  pivot_longer(cols = c(frac_CE, frac_MAE), names_to = "event_type", values_to = "fraction")

p1 <- clin_char_long %>% 
  mutate(event_type = str_replace_all(event_type, 
                                      c("frac_CE" = "Cardiac events", 
                                        "frac_MAE" = "Major arrhythmic events"))) %>% 
  ggplot(aes(x = gene, y = fraction, fill = gene)) + 
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(face = "italic",
                                   margin = margin(4, 0, 0, 0),
                                   hjust = 0.5),
        legend.position = "none") +
  facet_wrap(~event_type) +
  scale_fill_manual(values = selected_colors$gene) +
  labs(x = "",
       y = "Fraction of missense carriers with symptoms") +
  selected_theme +
  ggtitle("International Calmodulinopathy Registry (v2)")

p1

clin_char_df %>% 
  kable() %>% 
  kableExtra::kable_styling(., bootstrap_options = c("condensed", 
        "striped"), full_width = FALSE, font_size = 14)
```

##### Test Cardic events
```{r fig.width = 6, fig.height = 6, out.width = "100%", eval = TRUE}
test_CE <- chisq.test(clin_char_df %>% select(cardiac_events, free_from_CE))
test_CE

# test_CE$observed
# test_CE$expected

# row_totals <- rowSums(test_CE$observed)
# col_totals <- colSums(test_CE$observed)
# grand_total <- sum(test_CE$observed)
# 
# expected_manual <- outer(row_totals, col_totals) / grand_total
# expected_manual 
# 
# identical(expected_manual, test_CE$expected) # identical results
# 
# chi_sq <- sum((test_CE$observed - expected_manual)^2 / expected_manual) # chi squared statistic
# df <- (nrow(test_CE$observed) - 1) * (ncol(test_CE$observed) - 1) # degrees of freedom
# 1 - pchisq(chi_sq, df)

# fisher.test(clin_char_df %>% select(cardiac_events, free_from_CE))
```
<br><br>
  

### Any events (with subtypes)
```{r Any events (with subtypes), fig.width = 7, fig.height = 6, out.width = "100%"}
p1 <- clin_char_long %>% 
  filter(event_type != "frac_MAE") %>% 
  mutate(event_type = str_replace_all(event_type, 
                                      c("frac_CE" = "Cardiac events", 
                                        "frac_MAE" = "Major arrhythmic events")),
         numbers_MAE = paste0(MAE, "/", n_carriers),
         numbers_CE = paste0(cardiac_events, "/", n_carriers)) %>% 
  ggplot(aes(x = gene, y = fraction, fill = gene)) + 
  geom_bar(stat = "identity") +
  geom_text(aes(label = numbers_CE),
            position = position_stack(vjust = 1.05), # or vjust = 1.05 for above bars
            # color = "black", size = 8) + # settings for poster
            color = "black", size = 6) + # settings for graphical abstrac
            # color = "black", size = 4) + # settings for paper
   labs(x = "", 
        y = "Missense carriers with any cardiac event",
        title = "All carriers (n = 140)") +
  selected_theme +
  theme(axis.text.x = element_text(face = "italic",
                                   hjust = 0.5),
        legend.position = "none") + 
  scale_y_continuous(
    labels = scales::label_percent(), 
    breaks = seq(0, 1, by = 0.250),
    limits = c(0, 1))

clin_char_phenotyped <- data.frame(
  gene = cam_genes,
  n_carriers = c(18, 26, 30, 20, 11, 5),
  MAE = c(14, 13, 9, 10, 6, 4),
  CE = c(16, 16, 15, 20, 8, 5),
  subgroup = c("LQTS", "LQTS", "LQTS", "CPVT", "CPVT", "CPVT")) %>% 
  mutate(frac_MAE = MAE/n_carriers,
         frac_CE = CE/n_carriers)

clin_char_phenotyped_long <- clin_char_phenotyped %>%
  pivot_longer(cols = c(frac_CE, frac_MAE), names_to = "event_type", values_to = "fraction") %>% 
  mutate(subgroup = factor(subgroup, levels = c("LQTS", "CPVT")),
         # event_type = str_replace_all(event_type, 
         #                              c("frac_CE" = "Cardiac events", 
         #                                "frac_MAE" = "Major arrhythmic events")),
         numbers_MAE = paste0(MAE, "/", n_carriers),
         numbers_CE = paste0(CE, "/", n_carriers))


p2 <- clin_char_phenotyped_long %>% 
  filter(event_type != "frac_MAE") %>% 
  filter(subgroup == "LQTS") %>% 
  ggplot(aes(x = gene, y = fraction, fill = gene)) + 
  geom_bar(stat = "identity") +
  geom_text(aes(label = numbers_CE),
            position = position_stack(vjust = 1.1), # or vjust = 1.05 for above bars
            color = "black", size = 4) +
  selected_theme +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "none") +
  labs(x = "",
       y = "")

p3 <- clin_char_phenotyped_long %>% 
  filter(event_type != "frac_MAE") %>% 
  filter(subgroup == "CPVT") %>% 
  ggplot(aes(x = gene, y = fraction, fill = gene)) + 
  geom_bar(stat = "identity") +
  geom_text(aes(label = numbers_CE),
            position = position_stack(vjust = 1.08), # or vjust = 1.05 for above bars
            color = "black", size = 4)  +
  selected_theme +
  theme(axis.text.x = element_text(face = "italic",
                                   hjust = 0.5),
        legend.position = "none") +
  labs(x = "",
       y = "")

combined <- p2 + ggtitle("LQTS phenotype (n = 74)") +
  p3 + ggtitle("CPVT phenotype (n = 36)") + 
  plot_layout(ncol = 1, guides = "collect",  axes = "collect") &
  # labs(x = "",
  #      y = "") &
  scale_y_continuous(
    labels = scales::label_percent(), 
    breaks = seq(0, 1, by = 0.250)) 

p4 <- p1 +
  combined + 
  plot_layout(ncol = 2, guides = "collect",  axes = "collect") &
  plot_annotation(title = "International Calmodulinopathy Registry (v2)") & 
  scale_fill_manual(values = selected_colors$gene) 
# &
#   theme(legend.position = "none") 

p4
```
<br><br>

```{r Any events (with subtypes) 7_5, fig.width = 7, fig.height = 5, out.width = "100%"}
p4
```

```{r Any event 5_6, fig.width = 5, fig.height = 6, out.width = "100%"}
p1 <- p1 + 
  # ggtitle("International Calmodulinopathy Registry (v2)") +
  ggtitle("") +
  scale_fill_manual(values = selected_colors$gene) +
  theme(legend.position = "none") 
p1

selected_theme_poster <- theme(
  axis.text.y = element_text(size = 18),
  axis.text.x = element_text(size = 18),
  axis.title.y = element_text(size = 20), 
  axis.title.x = element_text(size = 20), 
  legend.text = element_text(size = 18),
  legend.title = element_text(size = 20),
  title = element_text(size = 18))

# p1 + selected_theme_poster
```
<br><br>

```{r Any event 6_5, fig.width = 6, fig.height = 5, out.width = "100%", eval = FALSE}
p1
```

```{r Any event 7_7, fig.width = 7, fig.height = 7, out.width = "100%"}
p1 + selected_theme_poster
```
<br><br>
  
#### Test subtypes
```{r Test subtypes, fig.width = 10, fig.height = 6, out.width = "100%"}
clin_char_phenotyped <- clin_char_phenotyped %>% 
  mutate(free_from_CE = n_carriers - CE,
         free_from_MAE = n_carriers - MAE)
# clin_char_phenotyped %>%
#   group_by(subgroup) %>% 
#   reframe(sum = sum(n_carriers))

t1 <- clin_char_phenotyped %>% 
  filter(subgroup == "LQTS") %>% 
  select(free_from_CE, CE)

t1
fisher.test(t1)

t2 <- clin_char_phenotyped %>% 
  filter(subgroup == "CPVT") %>% 
  select(free_from_CE, CE)
t1
fisher.test(t2)
```
<br><br>


#### Comparison any event and MAE
```{r Comparison any event and MAE, fig.width = 6, fig.height = 6, out.width = "100%", eval = FALSE}
p1 <- clin_char_long %>% 
  filter(event_type != "frac_MAE") %>% 
  mutate(event_type = str_replace_all(event_type, 
                                      c("frac_CE" = "Cardiac events", 
                                        "frac_MAE" = "Major arrhythmic events")),
         numbers_MAE = paste0(MAE, "/", n_carriers),
         numbers_CE = paste0(cardiac_events, "/", n_carriers)) %>% 
  ggplot(aes(x = gene, y = fraction, fill = gene)) + 
  geom_bar(stat = "identity") +
  geom_text(aes(label = numbers_CE),
            position = position_stack(vjust = 1.05), # or vjust = 1.05 for above bars
            color = "black", size = 5) 

p5 <- clin_char_long %>% 
  filter(event_type != "frac_CE") %>% 
  mutate(event_type = str_replace_all(event_type, 
                                      c("frac_CE" = "Cardiac events", 
                                        "frac_MAE" = "Major arrhythmic events")),
         numbers_MAE = paste0(MAE, "/", n_carriers),
         numbers_CE = paste0(cardiac_events, "/", n_carriers)) %>% 
  ggplot(aes(x = gene, y = fraction, fill = gene)) + 
  geom_bar(stat = "identity") +
  geom_text(aes(label = numbers_MAE),
            position = position_stack(vjust = 1.05), # or vjust = 1.05 for above bars
            color = "black", size = 5)

p1 + ggtitle("Any cardiac event") +
  p5 + ggtitle("Major arrhythmic event") + 
  plot_layout(ncol = 2, guides = "collect",  axes = "collect") &
  labs(x = "",
       y = "Symptoms among missense variant carriers") &
  scale_y_continuous(
    labels = scales::label_percent(), 
    breaks = seq(0, 1, by = 0.250),
    limits = c(0,1))  &
  selected_theme &
  scale_fill_manual(values = selected_colors$gene) &
  theme(axis.text.x = element_text(face = "italic",
                                   hjust = 0.5),
        legend.position = "none") 
```
<br><br>

### GnomAD variants
```{r}
v4 <- read_tsv(here(gnomAD_path, "gnomad.v4.1.constraint_metrics.tsv"))
v4_cam <- v4 %>% 
  filter(str_detect(transcript, "ENST") & mane_select == TRUE) %>% 
  filter(gene %in% cam_genes)

plot_oe <- function(subset_long) {
  ggplot(subset_long, aes(x = gene, y = oe, color = group)) +
    geom_point(position = position_dodge(width = 0.4), size = 3) +
    geom_linerange(aes(ymin = oe_lower, ymax = oe_upper),
                   position = position_dodge(width = 0.4), linewidth = 1) +
    theme(axis.text.x = element_text(face = "italic",
                                     # angle = 45,
                                     # hjust = 1,
                                     margin = margin(4, 0, 0, 0),
                                     hjust = 0.5)) +
    scale_color_manual(values = c("orange", "darkred", "#525252")) +
    labs(x = "",
         y = "Observed/expected",
         color = "Functional class \nof mutation") +
    geom_hline(yintercept = 1, linetype = "dashed", color = "black", linewidth = 1)
}

v4_cam_conf_scores <- left_join(
  v4_cam %>% 
    select(gene, mis.oe_ci.lower, lof.oe_ci.lower, syn.oe_ci.lower) %>% 
    pivot_longer(cols = -gene, names_to = "group", values_to = "oe_lower") %>% 
    mutate(group = str_remove(group, "_ci.lower")),
  v4_cam %>% 
    select(gene, mis.oe_ci.upper, lof.oe_ci.upper, syn.oe_ci.upper) %>% 
    pivot_longer(cols = -gene, names_to = "group", values_to = "oe_upper") %>% 
    mutate(group = str_remove(group, "_ci.upper")),
  by = c("gene", "group")) %>% 
  mutate(gene = factor(gene, levels = cam_genes))
  
v4_cam_conf_scores <- left_join(
  v4_cam_conf_scores,
  v4_cam %>% 
    select(gene, mis.oe, lof.oe, syn.oe) %>% 
    pivot_longer(cols = -gene, names_to = "group", values_to = "oe"),
  by = c("gene", "group")) %>% 
  mutate(group = str_replace_all(group, c(
    # "lof.oe" = "Loss-of-function",
    "lof.oe" = "Predicted \nloss-of-function",
                                          "mis.oe" = "Missense",
                                          "syn.oe" = "Synonymous")))
p1 <- v4_cam_conf_scores %>% 
  filter(group != "Synonymous") %>% 
  mutate(group = factor(group, levels = c("Predicted \nloss-of-function", "Missense"))) %>% 
  plot_oe + 
  selected_theme 
```

```{r GnomAD variants 5_5, fig.width = 5, fig.height = 5, out.width = "100%", eval = FALSE}
p1
```

```{r GnomAD variants 5_4, fig.width = 5, fig.height = 4, out.width = "100%"}
p1
```


```{r GnomAD variants 7_7, fig.width = 7, fig.height = 7, out.width = "100%"}
p1 
p1 + ggtitle("gnomAD (v4.1)") + selected_theme_poster
```
<br><br>

Scores are shown below:
```{r}
v4_cam_conf_scores %>% s_table
```
<br><br>
  


------------------------------------------------------------------------
```{r computetime, echo=FALSE}
time <- lubridate::date()
```
Report was finished `r time` <br>


```{r}
sessionInfo()
```


```{r rendering while keeping objects in global environment, echo = FALSE, eval=FALSE}
rmarkdown::render("scripts/variants_and_clinical_data.Rmd", clean = TRUE, output_dir = "reports/")
```
<br>
